---
title: "Project 4 Report"
author: "Joe Nagel"
date: ""
output: 
  pdf_document:
    number_sections: true
fontsize: 12pt
mainfont: Helvetica
geometry: margin=1in
header-includes:
  - \usepackage{float}
  - \usepackage{setspace}
  - \doublespacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(list(echo = FALSE, include = FALSE))
```

```{r}
# read in packages
library(tidyverse)
library(knitr)
library(kableExtra)
```

```{r}
# read in experiment results
params <- readRDS("../Data/simulation_parameters.rds")
sim <- (
    read_csv("../Data/simulation_results.csv", show_col_types = FALSE)
    %>% mutate(
        selection_method = factor(
            selection_method,
            levels = c(
                "p",
                "aic",
                "bic",
                "lasso_min",
                "elanet_min",
                "lasso_1se",
                "elanet_1se"
            )
        ),
        selected = TRUE
    )
)
```

```{r}
(
    expand_grid(
        n = params$n,
        rho = params$rho,
        replication = seq_len(params$replications),
        selection_method = c(
            "p",
            "aic",
            "bic",
            "lasso_min",
            "elanet_min",
            "lasso_1se",
            "elanet_1se"
        ),
        term = paste0("V", sprintf("%02d", 1:20))
    )
    %>% left_join(sim, by = join_by(n, rho, replication, selection_method, term))
    %>% mutate(
        .after = term,
        true_effect = recode_values(
            term,
            from = paste0("V", sprintf("%02d", 1:20)),
            to = params$beta
        ),
        p_value = replace_na(p_value, 1)
    )
    %>% mutate(across(c(estimate, conf_low, conf_high), ~replace_na(., 0)))
)
```


```{r}
(
    sim
    %>% summarise(
        .by = c(n, rho, selection_method, term),
        prop_selected = n()/params$replications
    )
    %>% arrange(term)
    %>% pivot_wider(
        names_from = term,
        values_from = prop_selected
    )
    %>% mutate(.keep = "unused", V06_V20 = rowMeans(across(V06:V20)))
    %>% arrange(n, selection_method, rho)
)
```

```{r}
# create confusion matrix
positive = paste0("V", sprintf("%02d", 1:5))
negative = paste0("V", sprintf("%02d", 6:20))
(
    sim
    %>% group_by(n, rho, selection_method, replication)
    %>% summarise(
        true_positive = all(positive %in% term),
        false_positive = any(negative %in% term),
        .groups = "drop_last"
    )
    %>% summarise(
        tpr = mean(true_positive), 
        tnr = 1 - mean(false_positive),
        .groups = "drop"
    )
    %>% pivot_longer(cols = c(tpr, tnr))
    %>% ggplot(aes(x = paste(n, rho, sep = "\n"), y = selection_method, fill = value))
        + geom_tile(color = "white")
        + geom_text(
            aes(
                label = round(value, 2),
                color = value < 0.5
            ),
            size = 3
        )
        + scale_color_manual(
            values = c("TRUE" = "white", "FALSE" = "black"),
            guide = "none"
        )
        + facet_wrap(
            ~ name,
            labeller = labeller(
                name = c(
                    tpr = "True Positive Rate",
                    tnr = "True Negative Rate"
                )
            )
        )
        + scale_y_discrete(limits = rev)
        + scale_fill_viridis_c(name = "Rate", limits = c(0, 1))
        + theme_minimal()
        + labs(
            x = "Sample Size\nCorrelation",
            y = "Selection Method"
        )
)
```


# Introduction

In statistical analysis, defining a statistical model often requires choosing predictors and/or covariates from a large set of variables. In controlled settings such as clinical trials, explanatory variables, covariates, and the corresponding final model can be defined a priori. However, in cases where such a procedure is not possible, variable selection, the process of choosing a set of variables relevant to the outcome of interest, is often required. The most straightforward method for variable selection is to fit all possible combinations of the candidate predictors and choose a final model based on a measure of model fit; common measures include quantifications of predictive capability like mean square error, likelihood based measures such as Akaike Information Criterion (AIC) and Bayesian Information Criterion (BIC), and statistical comparisons of model fits with F-tests. However, this approach is only feasible for analyses with a relatively small pool of candidate predictors. Backward selection is a common approach that overcomes this problem by fitting a model with all candidates comparing it to reduced models excluding one predictor at a time. The model with the best fit is selected and this process is repeated until no predictors remain or the model fit cannot be improved by removing a predictor. Another approach is to place an L1 penalty on the vector of regression coefficients ($\beta$). The results in the shrinkage of some coefficients to zero, effectively removing the corresponding variable from the model. This can be done using only an L1 penalty (lasso regression) or a mix of L1 and L2 penalties (elastic net regression). While these approaches often result in more parsimonious models, they can overfit to noise in the dataset and do not account for the selection procedure during model inference. This analysis will explore how these approaches perform in an array of simulated scenarios. Specifically, we will compare the performance of each method while varying sample size and the correlation between variables. Performance will be measured by assessing each methods ability to first capture variables with a true relationship with the outcome and non-important variables and then produce valid inference for the selected variables.

# Preliminary Methods

The purpose of this analysis is to explore the performance of different variable selection methods across differing sample sizes and correlations. Specifically, we will create simulated datasets with 20 variables sampled from a multivariate normal distribution with an exchangeable correlation structure. Five of the variables will be associated with the outcome with varying strength ranging from 1/6 to 5/6 with equal spacing. Standard normal noise will be added to the response. Datasets will be simulated for two sample sizes ($n=250,500$) and three correlations ($\rho=0, 0.35, 0.7$) resulting in six combinations. Each dataset will be replicated 10,000 times to ensure the accuracy of the estimated parameters. Using the simulated data, we will compare the performance of five different methods for variable selection: Backward selection based on p-values (from F-tests), AIC, and BIC as well as lasso regression and elastic net regression (with mixing parameter $\alpha = 0.5$). All regression models will be fit without an intercept term. P-value based Backward selection was performed with $p \ge 0.15$ as the threshold for excluding a parameter. For the lasso and elastic net approaches, we will use 10-fold cross validation to select the penalty parameter, $\lambda$, based on the average out-of-fold mean square error (MSE) and will compare the $\lambda$ that achieved the minimum MSE with the largest $\lambda$ within one standard error of the minimum. For the lasso and elastic net approaches, a simple linear regression model will be fit including all variables with non-zero coefficients. Selection performance will be assessed with the true positive rate, how often the method selected all five variables associated with the outcome, and true negative rate, how often the method did not select any of the variables not associated with the outcome. Additionally, we will investigate the rate of Type I and Type II errors, the bias of the parameter estimates, and the coverage of Wald-type 95% confidence intervals from each model. Any variables not selected will be assumed to be non-significant with a point estimate of zero with no error. All simulations were conducted using R (version 4.5.1). Random data generation and fold assignment will be controlled using a fixed seed for reproducibility.

<!-- # Results -->

<!-- # Discussion and Conclusion -->

# Reproducible Research

## GitHub Link

`https://github.com/nageljo/BIOS6624`
