---
title: "Research Strategy and Sample Size Estimation"
author: "Joe Nagel"
date: ""
output: 
  pdf_document:
    number_sections: false
fontsize: 12pt
mainfont: Helvetica
geometry: margin=1in
header-includes:
  - \usepackage{float}
  - \usepackage{setspace}
  - \doublespacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE)
```

```{r}
library(tidyverse)
library(powertools)
library(knitr)
library(kableExtra)
```

```{r}
# read in preliminary data
prelim <- read_csv('../Data/PrelimData.csv', show_col_types = FALSE)
```

```{r}
names(prelim)
```

```{r}
cor(prelim)
```


```{r}
# fit preliminary models
cvlt_mod <- lm(CVLT_CNG3 ~ IL_6, data = prelim)
cort_mod <- lm(CORT_CNG3 ~ IL_6 + MCP_1, data = prelim)
```

```{r}
cvlt_R2 <- summary(cvlt_mod)$r.squared
summary(cvlt_mod)
```

```{r}
cort_R2 <- summary(cort_mod)$r.squared
summary(cort_mod)
```

```{r}
# create table for Aim 1 omnibus power
# define parameters
N = 175
p = 4
alpha = 0.025
power = NULL

# create sequence of R squared values and calculate power
table_aim1_omni = (
    tibble(
        Rsq = c(0.01, seq(from = 0.025, to = 0.15, by = 0.025)),
        power = map_dbl(
            .x = Rsq,
            .f = \(Rsq) mlrF.overall(
                N = N,
                p = p,
                Rsq = Rsq,
                alpha = alpha,
                power = power
            )
        )
    )
    %>% round(3)
    %>% kable(
        format = "latex",
        booktabs = TRUE,
        escape = FALSE,
        linesep = "",
        align = "cc",
        col.names = c("$R^2$", "Power"),
        caption = "Power to Detect Overall Inflammation Marker Effects ($\\alpha = 0.025$)"
    )
    %>% kable_styling(latex_options = c("HOLD_position"))
)
```

```{r}
# create table for Aim 1 individual predictor power
# define parameters
N = 175
rho0 = 0
alpha = 0.025
power = NULL

# create sequence of R squared values and calculate power
table_aim1_indv = (
    tibble(
        correlation = seq(from = 0.15, to = 0.5, by = 0.05),
        power = map_dbl(
            .x = correlation,
            .f = \(rhoA) corr.1samp(
                N = N,
                rho0 = rho0,
                rhoA = rhoA,
                alpha = alpha,
                power = power
            )
        )
    )
    %>% round(3)
    %>% kable(
        format = "latex",
        booktabs = TRUE,
        escape = FALSE,
        linesep = "",
        align = "cc",
        col.names = c("$\\rho$", "Power"),
        caption = "Power to Detect Individual Inflammation Marker Effects ($\\alpha = 0.025$)"
    )
    %>% kable_styling(latex_options = c("HOLD_position"))
)
```

```{r}
# create table for Aim 2 power 
N = 175
alpha = 0.025
power = NULL
rho1 = 0.1

table_aim2 = (
    expand_grid(
        prop_pos = seq(from = 0.1, to = 0.9, by = 0.2),
        rho2 = seq(from = 0.2, to = 0.9, by = 0.1)
    )
    %>% mutate(
        prop_neg = 1 - prop_pos,
        n1 = ceiling(N*prop_neg),
        n2 = N - n1,
        n.ratio = n2/n1,
        delta = rho2 - rho1,
        power = pmap_dbl(
            .l = list(n1 = n1, n.ratio = n.ratio, rho2 = rho2),
            .f = \(n1, n.ratio, rho2) corr.2samp(
                n1 = n1,
                n.ratio = n.ratio,
                rho1 = rho1,
                rho2 = rho2,
                alpha = alpha,
                power = power
            )
        )
        
    )
    %>% mutate(.keep = "none", delta, power, percent_pos = prop_pos)
    %>% pivot_wider(
        names_from = percent_pos,
        values_from = power
    )
    %>% round(3)
    %>% kable(
        format = "latex",
        booktabs = TRUE,
        escape = FALSE,
        linesep = "",
        align = "cccccc",
        col.names = c(
            "$\\Delta \\rho$",
            paste0(seq(10, 90, by = 20), "\\%")
        ),
        caption = "Power to Detect Amyloid Deposition Inflammation Interaction Effects ($\\alpha = 0.025$)"
    )
    %>% kable_styling(latex_options = c("HOLD_position"))
    %>% add_header_above(c(" " = 1, "Percent with Amyloid Deposition" = 5))
)
```

#### Aim 1 Research Strategy
[including how you go from the data you will get from the investigator to your final data set and how you will assess/investigate any impact of the differences between the whole dataset and the one you would use for analysis]

[this could use improvement]
Prior to analysis, all variables will be assessed for informative missingness, normality, and outliers. All six outcomes of interest for this aim are continuous and include AD-signature cortical thickness and the five cognitive measures enumerated in *Grant Table 3*, capturing recall/recognition, pattern separation, and long term memory consolidation. We will use two cytokines (IL-6 and TNF-alpha) and two chemokines (MCP-1 and eotaxin-1) as our measures of peripheral inflammation. To investigate the relationship between the baseline measures of peripheral inflammation and the outcomes of interest (**Hypothesis 1a**), we will model the baseline minus Year‑1 change in each outcome as a function of the baseline inflammation measures and the baseline value of the outcome using regression analysis. We will first evaluate the joint significance of the inflammation measures using an omnibus test. If the omnibus test is significant, we will then examine associations between the inflammation markers and the outcome using standardized estimates and confidence intervals. Preliminary data suggests that the inflammation measures may be highly correlated. If this causes high variance inflation, we will model each measure separately. To determine if increases in measures of inflammation are associated with worse outcomes (**Hypothesis 1b**), we will refit the previous models using the baseline minus Year-1 change in inflammation measures in place of the baseline inflammation measures and follow the same analysis procedure. All models fit to address **Hypotheses 1a** and **1b** will control for age and sex. We will also consider controlling for other variables potentially associated with inflammation: cardiovascular risk markers (BMI, blood pressure, and history of hypercholesterolemia), NSAID use, and immune-related health conditions (Crohn's disease, irritable bowel syndrome, and arthritis). Specifically, we will prioritize variables with the potential to function as mediators, confounders, or precision variables, based on their correlations with outcomes and/or measures of inflammation.

We will control our Type 1 error for each hypothesis by accounting for testing each of two main outcomes, AD-signature cortical thickness and cognitive function. We expect that the cognitive measure will be correlated and preliminary data suggests that the measures of inflammation are correlation, so we will not adjust for multiple comparisons within these groups.

[this needs a full rewrite]
We will perform a power analysis for both **Hypotheses 1a** and **1b** by simplifying each regression into a correlation between an inflammation measure and an outcome. We will define both expected and clinically meaningful correlations and investigate the power to detect the relationships at the current sample size of 175. We will also calculate power for additional sample sizes to establish a complete characterization of the required sample size. Because this approach does not account for the additional covariates in the analysis, in the absense of confounders [True.  Since there could be confounders, but also precision, could instead point out that you expect both and so did not adjust the expected correlation accordingly.], it will produce conservative estimates.

#### Aim 2 Research Strategy

This Aim involves the same inflammation measures and outcomes as **Aim 1** in addition to a continuous measure of amyloid deposition. To investigate the interaction between amyloid deposition and peripheral inflammation markers (**Hypothesis 2**), we will refit the models from **Hypothesis 1b** with the addition of a main effect of amyloid deposition and an interaction term between the change in the inflammation markers and amyloid deposition. We will only refit models that demonstrated a significant relationship between the inflammation markers and outcome. If no models demonstrated a significant relationship, all six models will be refit. This will allow us to determine whether declines in memory and AD‑signature cortical thickness are greatest when both amyloid deposition is high and peripheral inflammation is elevated. These models will adjust for age, sex, and any additional measures related to inflammation found through the exploratory process described in **Aim 1**.
[One other detail to consider is a sentence or two about how you will describe any significant interactions - how will you report to make it meaningful beyond the p-value?]

[this needs a full rewrite]
Consistent with Aim 1, we will estimate power by simplifying the interaction effect of interest to an equivalent correlation. We will calculate the power to detect expected and clinically meaningful correlations across a range of sample sizes, including the current target sample of 175.


```{r, include=TRUE}
table_aim1_omni
```

```{r, include=TRUE}
table_aim1_indv
```

```{r, include=TRUE}
table_aim2
```