---
title: "Risk Factors and Probability of Stroke"
author: "Joe Nagel"
date: ""
output: 
  pdf_document:
    number_sections: true
fontsize: 12pt
mainfont: Helvetica
geometry: margin=1in
header-includes:
  - \usepackage{float}
  - \usepackage{setspace}
  - \doublespacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(list(echo = FALSE, include = FALSE))
```

```{r}
# read in packages
library(tidyverse)
library(naniar)
library(janitor)
library(knitr)
library(kableExtra)
library(gtsummary)
TEN_YEARS = 365.25*10
```

```{r}
(
    read_csv(
        '../Data/frmgham2.csv',
        show_col_types = FALSE,
        name_repair = "unique_quiet"
    )
    %>% clean_names()
    %>% mutate(
        .before = randid,
        stroke_in_followup = if_else(timestrk <= TEN_YEARS, stroke, 0),
        stroke_time = pmin(timestrk, TEN_YEARS)
    )
) -> fram
```

```{r}
naniar::vis_miss(fram)
```

```{r}
names(fram)
```

```{r}
sample_n(fram, size = 10)
```

```{r}
(
    fram
    %>% tbl_summary(
        by = sex,
        statistic = list(
            all_continuous() ~ "{mean} ({sd})",
            all_categorical() ~ "{n} ({p}%)"
        ),
        missing = "if_any",
        missing_text = "Missing/Improbable, n (%)",
        missing_stat = "{N_miss} ({p_miss}%)",
        include = c(
            totchol,
            age,
            sysbp,
            cursmoke,
            bmi,
            diabetes,
        )
    )
)
```



Do the analysis separately for males and female (the models can be different between sexes if there are different relationships)

Risk factors (baseline only)
    include: age, diabetes, blood pressure (systolic)
    several risk factors not sure about: coronary heart disease, blood pressure medication, smoking status (current Y/N), total cholesterol, BMI
    
Exclusion criteria:
    only those who have not already had a stroke
    check for outliers
    
only interested in strokes

We can treat events other than stroke as censored 

we need descriptives of how predictors vary over time (we are not doing a time varying analysis)

Only check time varying for diabetes and blood pressure (use descriptives and interpret how a time varying analysis may change results)

censor events outside of the first 10 years of followup





# Reproducible Research

## GitHub Link

`https://github.com/nageljo/BIOS6624`

## Dataset Filepath

`C:\Users\josep\OneDrive\AdvancedData\BIOS6624\Project3\Data\frmgham2.csv`