---
title: "Risk Factors and Probability of Stroke"
author: "Joe Nagel"
date: ""
output: 
  pdf_document:
    number_sections: true
fontsize: 12pt
mainfont: Helvetica
geometry: margin=1in
header-includes:
  - \usepackage{float}
  - \usepackage{setspace}
  - \doublespacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(list(echo = FALSE, include = FALSE))
```

```{r}
# read in packages
library(tidyverse)
library(patchwork)
library(naniar)
library(janitor)
library(knitr)
library(kableExtra)
library(gtsummary)
library(survival)
library(survminer)
library(broom)
TEN_YEARS = 365.25*10
```

```{r}
(
    read_csv(
        '../Data/frmgham2.csv',
        show_col_types = FALSE,
        name_repair = "unique_quiet"
    )
    %>% clean_names()
    %>% mutate(
        .before = randid,
        stroke_in_followup = if_else(0 < timestrk & timestrk <= TEN_YEARS, stroke, 0),
        stroke_time = pmin(timestrk, TEN_YEARS)/365.25,
        sex = factor(sex, levels = 1:2, labels = c("Male", "Female")),
        bmi_cat = cut(
            bmi,
            breaks = c(-Inf, 25, 30, Inf),
            labels = c("Normal", "Overweight", "Obese")
        ),
        chol_cat = cut(
            totchol,
            breaks = c(-Inf, 240, Inf),
            labels = c("Normal", "High")
        )
    )
) -> fram
```

```{r}
# get total number of patients
fram$randid %>% unique() %>% length()
```

```{r}
# check for missingness
# naniar::vis_miss(fram)
```

```{r}
# create analysis dataset
fram_ana <- (
    fram
    # get baseline covariates and remove subjects who had a stroke before baseline
    %>% filter(period == 1 & prevstrk == 0)
    %>% select(randid, sex, stroke_in_followup, stroke_time, age, sysbp, totchol,
               bmi, diabetes, prevchd, bpmeds, cursmoke, bmi_cat, chol_cat)
)
```

```{r}
# create table 1
table_one <- (
    fram_ana
    %>% mutate(stroke_time = stroke_time)
    %>% tbl_summary(
        by = sex,
        statistic = list(
            all_continuous() ~ "{mean} ({sd})",
            all_categorical() ~ "{n} ({p}%)"
        ),
        missing = "ifany",
        missing_text = "Missing, n (%)",
        missing_stat = "{N_miss} ({p_miss}%)",
        include = c(
            stroke_in_followup,
            stroke_time,
            age,
            sysbp,
            totchol,
            bmi,
            diabetes,
            prevchd,
            bpmeds,
            cursmoke
        ),
        label = list(
            stroke_in_followup ~ "Stroke During Follow-up, n (%)",
            stroke_time ~ "Follow-up Time (Years), mean (SD)",
            age ~ "Age, mean (SD)",
            sysbp ~ "Systolic Blood Pressure, mean (SD)",
            totchol ~ "Total Cholesterol, mean (SD)",
            bmi ~ "BMI, mean (SD)",
            diabetes ~ "Diabetes, n (%)",
            prevchd ~ "Coronary Heart Disease, n (%)",
            bpmeds ~ "Blood Pressure Mediation, n (%)",
            cursmoke ~ "Current Smoker, n (%)"
        ),
    )
    %>% add_overall(last = TRUE)
    %>% modify_footnote(everything() ~ NA)
    %>% as_kable_extra(
        booktabs = TRUE,
        linesep = "",
        caption = "Descriptive Statistics of Events, Follow-up, and Baseline Risk Factors",
        escape = FALSE
    )
    %>% kable_styling(latex_options = c("scale_down", "HOLD_position"))
)
```

```{r}
plot_km_curves <- (
    fram_ana
    %>% mutate(
        .keep = "none",
        stroke_time,
        stroke_in_followup,
        sex,
        `BP Meds` = factor(bpmeds, levels = 0:1, labels = c("No", "Yes")),
        `Current Smoker` = factor(cursmoke, levels = 0:1, labels = c("No", "Yes")),
        `Prev CHD` = factor(prevchd, levels = 0:1, labels = c("No", "Yes")),
        Cholesterol = chol_cat,
        BMI = bmi_cat
    )
    %>% drop_na()
    %>% pivot_longer(
        cols = !c(stroke_time, stroke_in_followup, sex),
        names_to = "variable",
        values_to = "value"
    )
    %>% group_by(sex, variable, value)
    %>% group_modify(~{
        fit <- survfit(Surv(stroke_time, stroke_in_followup) ~ 1, data = .x)
        tibble(time = fit$time, surv = fit$surv)
    })
    %>% ungroup()
    %>% ggplot(., aes(x = time, y = surv, color = value))
        + geom_step(linewidth = 0.9)
        + facet_grid(sex ~ variable)
        + labs(
            x = "Time (Years)",
            y = "Survival probability",
            color = "Group"
        )
        + scale_x_continuous(breaks = c(0, 5, 10))
        + theme_minimal()
)
```

```{r}
# fit Cox PH models
full_formula = Surv(stroke_time, stroke_in_followup) ~ age + diabetes + sysbp + totchol + bmi + prevchd + bpmeds + cursmoke
base_formula = Surv(stroke_time, stroke_in_followup) ~ age + diabetes + sysbp

coxph_full_male <- coxph(
    full_formula,
    data = fram_ana %>% filter(sex == "Male") %>% drop_na()
)

coxph_full_female <- coxph(
    full_formula,
    data = fram_ana %>% filter(sex == "Female") %>% drop_na()
)
```

```{r}
# run backwards selection
step(
  coxph_full_male,
  scope = list(
    lower = base_formula,
    upper = full_formula
  ),
  direction = "backward"
)
```

```{r}
step(
  coxph_full_female,
  scope = list(
    lower = base_formula,
    upper = full_formula
  ),
  direction = "backward"
)
```

```{r}
coxph_final_male <- coxph(
    update(base_formula, . ~ . + cursmoke),
    data = fram_ana %>% filter(sex == "Male") %>% drop_na()
)

coxph_final_female <- coxph(
    update(base_formula, . ~ . + cursmoke),
    data = fram_ana %>% filter(sex == "Female") %>% drop_na()
)
```

```{r}
table_model_coef <- (
    imap(
        list(male = coxph_final_male, female = coxph_final_female),
        ~ (
          tidy(.x, conf.int = TRUE, exponentiate = TRUE)
          %>% mutate(sex = .y, .before = 1)
        )
    )
    %>% list_rbind()
    %>% mutate(
        .keep = "none",
        term,
        sex,
        hazard_ratio = sprintf("%.3f", estimate),
        ci = sprintf("%.3f, %.3f", conf.low, conf.high),
        p_value = format.pval(p.value, digits = 2, eps = 0.001)
    )
    %>% pivot_wider(
        names_from = sex,
        values_from = c(hazard_ratio, ci, p_value)
    )
    %>% mutate(
        .keep = "unused",
        term = recode_values(
            term,
            "age" ~ "Age",
            "diabetes" ~ "Diabetes",
            "sysbp" ~ "Systolic Blood Pressure",
            "cursmoke" ~ "Smoking Status"
        )
    )
    %>% relocate(term,
                 hazard_ratio_male, ci_male, p_value_male,
                 hazard_ratio_female, ci_female, p_value_female)
    %>% kable(
        format = "latex",
        booktabs = TRUE,
        linesep  = "",
        caption  = "Cox Proportional Hazards Model Estimates",
        col.names = c("Coefficient",
                      "Hazard Ratio", "95% CI", "p-value",
                      "Hazard Ratio", "95% CI", "p-value"),
        align = "ccccc"
    )
    %>% row_spec(0, bold = TRUE)
    %>% add_header_above(c(" " = 1, "Male" = 3, "Female" = 3), bold = TRUE)
    %>% kable_styling(latex_options = c("scale_down", "HOLD_position"))
)
```



```{r}
coxph_res_male <- cox.zph(coxph_final_male)
coxph_res_female <- cox.zph(coxph_final_female)
```

```{r}
coxph_res_male$table
```

```{r}
coxph_res_female$table
```

```{r}
titles = c("Age (Years)", "Diabetes Status", "Systolic BP", "Smoking Status")

(
    ggcoxzph(
        coxph_res_male,
        se = FALSE
    )
    %>% map2(
        titles,
        \(plot, title) {(
            plot
            + ylab("Schoenfeld Residuals")
            + ggtitle(title)
        )}
    )
    %>% wrap_plots(res_plots_male, ncol = 2)
        + plot_layout(axis_titles = "collect")
)
```

```{r}
(
    ggcoxzph(
        coxph_res_female,
        se = FALSE
    )
    %>% map2(
        titles,
        \(plot, title) {(
            plot
            + ylab("Schoenfeld Residuals")
            + ggtitle(title)
        )}
    )
    %>% wrap_plots(res_plots_male, ncol = 2)
        + plot_layout(axis_titles = "collect")
)
```

```{r}
# get mean blood pressure values across the population
population_bp_means = (
    fram_ana
    %>% summarise(
        .by = sex,
        sysbp = mean(sysbp)
    )
)
```


```{r}
compute_year10_risk <- function(model, newdata) {
    fit <- survfit(model, newdata)
    est <- summary(fit, times = 10)
    
    year10_risk <- tibble(
        risk = 1 - drop(est$surv),
        lower_cl = 1 - drop(est$upper),
        upper_cl = 1 - drop(est$lower)
    )
    
    return(year10_risk)
}
```


```{r}
# create risk profiles
risk_profiles <- (
    tribble(
        ~sex,      ~profile,                ~diabetes, ~cursmoke,
        "Male",   "Baseline",               0,         0,
        "Male",   "High BP",                0,         0,
        "Male",   "Diabetes",               1,         0,
        "Male",   "High BP + Diabetes",     1,         0,
        "Male",   "Smoker",                 0,         1,
        "Female", "Baseline",               0,         0,
        "Female", "High BP",                0,         0,
        "Female", "Diabetes",               1,         0,
        "Female", "High BP + Diabetes",     1,         0,
        "Female", "Smoker",                 0,         1
    )
    %>% left_join(population_bp_means, by = "sex")
    %>% mutate(sysbp = if_else(str_detect(profile, "High BP"), 160, sysbp))
    %>% crossing(age = c(40, 50, 60))
    %>% group_by(sex)
    %>% group_modify(~{
        model <- if (.y$sex[[1]] == "Male") coxph_final_male else coxph_final_female
        bind_cols(.x, compute_year10_risk(model, newdata = .x))
    })
    %>% ungroup()
)
```

```{r}
# create table of risk probabilities for males
table_risk_male <- (
    risk_profiles
    %>% filter(sex == "Male")
    %>% mutate(
        .keep = "none",
        `Risk Profile` = profile,
        age,
        risk_ci = sprintf("%03.1f (%03.1f, %03.1f)", 100*risk, 100*lower_cl, 100*upper_cl)
    )
    %>% pivot_wider(names_from = age, values_from = risk_ci)
    %>% kable(
        format = "latex",
        booktabs = TRUE,
        linesep  = "",
        caption  = "10-Year Probability of Stroke for Males",
        align = "cccc"
    )
    %>% row_spec(0, bold = TRUE)
    %>% add_header_above(c(" " = 1, "Age (Years)" = 3), bold = TRUE)
    %>% kable_styling(latex_options = c("scale_down", "HOLD_position"))
    %>% footnote(general = "Values reported as % Chance (95% CI)")
)
```

```{r}
# create table of risk probabilities for females
table_risk_female <- (
    risk_profiles
    %>% filter(sex == "Female")
    %>% mutate(
        .keep = "none",
        `Risk Profile` = profile,
        age,
        risk_ci = sprintf("%03.1f (%03.1f, %03.1f)", 100*risk, 100*lower_cl, 100*upper_cl)
    )
    %>% pivot_wider(names_from = age, values_from = risk_ci)
    %>% kable(
        format = "latex",
        booktabs = TRUE,
        linesep  = "",
        caption  = "10-Year Probability of Stroke for Females",
        align = "cccc"
    )
    %>% row_spec(0, bold = TRUE)
    %>% add_header_above(c(" " = 1, "Age (Years)" = 3), bold = TRUE)
    %>% kable_styling(latex_options = c("scale_down", "HOLD_position"))
    %>% footnote(general = "Values reported as % Chance (95% CI)")
)
```

```{r}
table_time_varying <- (
    fram
    %>% filter(prevstrk == 0)
    %>% summarise(
        .by = c(period, sex),
        mean_sysbp = mean(sysbp),
        sd_sysbp = sd(sysbp),
        n_diabetes = sum(diabetes),
        percent_diabetes = mean(diabetes)*100,
        n = n()
    )
    %>% mutate(
        .keep = "none",
        period,
        Sex = sex,
        `Systolic Blood Pressure, mean (SD)` = sprintf("%.1f (%.1f)", mean_sysbp, sd_sysbp),
        `Diabetes, n (%)` = sprintf("%i (%.1f)", n_diabetes, percent_diabetes),
        `Number of Observations` = as.character(n)
    )
    %>% pivot_longer(cols = !c(period, Sex), names_to = "Characteristic")
    %>% mutate(
        Characteristic = factor(
            Characteristic,
            levels = c(
                "Number of Observations",
                "Systolic Blood Pressure, mean (SD)",
                "Diabetes, n (%)")
        )
    )
    %>% relocate(Characteristic, Sex)
    %>% arrange(Characteristic, Sex, period)
    %>% pivot_wider(names_from = period, values_from = value)
    %>% kable(
        format = "latex",
        booktabs = TRUE,
        linesep  = "",
        caption  = "Summary of Blood Pressure and Diabetes Status by Clinical Exam",
        align = "ccccc"
    )
    %>% row_spec(0, bold = TRUE)
    %>% add_header_above(header = c(" " = 2, "Collection Period" = 3), bold = TRUE)
    %>% kable_styling(latex_options = c("scale_down", "HOLD_position"))
    %>% collapse_rows(columns = 1)
)
```


# Introduction

Strokes are medical emergencies caused by restricted or blocked blood flow to the brain. They are often caused by blood clots or tears in the blood vessels that supply blood to the brain. This analysis investigates the factors that contribute to the 10-year risk of stroke in adults. The data come from the Framingham Heart Study, a prospective study of 5,209 subjects living in Framingham, Massachusetts which began in 1948. The subset used in this analysis includes demographic information as well as risk factors and biological markers of heart disease which were collected during 3 clinical exams. Each subject was also followed for 24 years for medical events related to heart disease: Angina Pectoris, Myocardial Infarction, Atherothrombotic Infarction, Cerebral Hemorrhage (Stroke), and death. The primary goal of this analysis is to identify risk factors of stroke and characterize the 10-year probability of stroke across specific risk profiles. Specifically, we will investigate the relationship between stroke and several potential risk factors: coronary heart disease (CHD), blood pressure medication, smoking status, total cholesterol, and body mass index (BMI). We will then model the 10-year probability of stroke separately for males and females as a function of baseline age, diabetes status, blood pressure and any additional risk factors we have identified. We will additionally investigate how the risk factors vary over time and any implications that variability might have for the results of the analysis.

<!-- # Preliminary Methods -->

<!-- All subjects with a history of stroke prior to baseline will be excluded from this analysis. The primary outcome is stroke within a 10-year follow-up period. Participants experiencing non-stroke events (e.g., Angina Pectoris, Myocardial Infarction, death) or remaining stroke-free beyond 10 years will be censored. Following an assessment for outliers and missing data, a complete case analysis will be performed, with baseline characteristics and events summarized by sex in Table 1. Cox Proportional Hazards models will be created separately for males and females. Age, diabetes status, and blood pressure will be maintained as fixed covariates in all models. We will then use AIC-based backward selection to evaluate coronary heart disease, BP medication, smoking status, total cholesterol, and BMI as potential risk factors of stroke. The proportional hazards model assumption will be validated using Kaplan-Meier survival curves and Grambsch-Therneau tests of Schoenfeld residuals. The fitted models will be used to estimate the 10-year probability of stroke across five distinct risk profiles. We will establish a baseline profile defined as an individual who is a non-smoker, does not have diabetes or CHD, and does not use BP medication, with total cholesterol, BMI, and blood pressure set at population averages. We will then compare this baseline to profiles reflecting individuals with high blood pressure (BP=160), diabetes, both high blood pressure and diabetes, and those classified as obese (BMI=30). Finally, an exploratory analysis of the longitudinal changes in diabetes and blood pressure will be conducted to evaluate the necessity of a time-varying covariate modeling approach. -->

# Methods

The analysis population included all participants in the Framingham cohort with no history of stroke at baseline; individuals with a prior stroke were excluded. Follow-up time was truncated at 10 years, and any observation time beyond 10 years was treated as censored at 10 years. For participants who experienced non-fatal, non-stroke events within the first 10 years (e.g., angina pectoris or myocardial infarction), follow-up continued, and fatal events were treated as censoring events. The primary risk factors of interest in this analysis were CHD, blood pressure medication, smoking status, total cholesterol, and BMI. All analysis variables were evaluated for missingness and outliers. A table of descriptive statistics of the analysis dataset was created. Additionally, all analyses were stratified by sex.

Kaplan–Meier curves were created to descriptively assess stroke-free survival across the risk factors of interest. Next, sex-stratified Cox proportional hazards models were fit, each containing the full set of candidate risk factors and adjusting for age, diabetes status, and systolic blood pressure. All models only included covariates measured at the first clinical exam (baseline). An AIC-based backward selection procedure was applied, with age, diabetes status, and systolic blood pressure retained as forced covariates. To improve interpretability and facilitate comparison by sex, the final male and female models retained any risk factor that was statistically significant in either sex. The proportional hazards assumption for the final models was checked using a combination of visualization and statistical testing of the Schoenfeld Residuals.

Final models were used to investigate the probability of stroke for a set of five distinct risk profiles. The first profile established a baseline risk and was defined as an individual who is a non-smoker, does not have diabetes or CHD, and does not use BP medication, with total cholesterol, BMI, and blood pressure set at population averages. This baseline was then compared to profiles reflecting individuals with high blood pressure (BP=160), diabetes, both high blood pressure and diabetes, and current smokers. The 10-year probability of stroke was calculated for each profile at ages 40, 50 and 60. Finally, an exploratory analysis examined changes in risk factors across clinical examinations, focusing on variability in systolic blood pressure and diabetes status over time. This assessment was used to evaluate potential limitations of relying solely on baseline measurements and to inform whether a time-varying covariate modeling approach might be warranted.

All analysis were performed using a type one error rate ($\alpha$) of 0.05. All models and figures were created with R version 4.5.1. ChatGPT version 5.2 was used to debug code, outline the report, and copy-edit the report.

# Results

Of the original 4434 subjects, 32 had a history of stroke and were excluded from the analysis. The remaining 4402 comprised the analysis dataset. *Table 1* summarizes the baseline covariates, events, and follow-up time for these subjects. Smoking was more prevalent among men than women, with 61% of men and 40% of women reporting current smoking at baseline. For the complete case analysis, 127 subjects with missing BMI, total cholesterol, or blood pressure medication use values were dropped, resulting in a final sample size of 4275 subjects. High values of systolic blood pressure ($295 \; mmHg$), total cholesterol ($696 \; mg/dL$), and BMI ($56 \; kg/m^2$) were observed. However, these values were biologically plausible and were retained for analysis.

```{r, include=TRUE}
table_one
```

The marginal stroke-free survival for each candidate risk factor is visualized in *Figure 1*. Blood pressure medication used and prevalent CHD appear to have the strongest relationship with stroke for both men and women. However, very few subjects used blood pressure medication (3.1%) or had CHD diagnosed before the study (4.2%). Additionally of note, the relationship between high cholesterol ($> 240 \; mg/dL$) and stroke appeared to be stronger among women than men.

```{r, include=TRUE, fig.cap="Kaplain-Meier Curves by Sex for Candidate Risk Factors"}
plot_km_curves
```

After AIC-based backwards selection, the model for males retained only smoking status from the candidate predictors and the model for females only included the fixed covariates. Following the procedure defined above, the final models for both males and females included age, diabetes status, systolic blood pressure, and smoking status. The model coefficients on the hazard ratio scale are shown in *Table 2*. For both males and female, an increase in age and systolic blood pressure were significantly related to an increased hazard of stroke (all p < 0.001). Among males, age (p < 0.001), diabetes (p < 0.001), systolic blood pressure (p < 0.001) and smoking status (p = 0.038) were all significantly related to the hazard of stroke. Among females, however, only age (p < 0.001) and systolic blood pressure (p < 0.001) were significantly related to the hazard of stroke, while diabetes (p = 0.254) and smoking status (p = 0.207) were not.

```{r, include=TRUE}
table_model_coef
```

The final sex-specific models were used to estimate the 10-year probability of stroke for each of the five risk profiles. These estimates are presented for males in *Table 3* and for females in *Table 4*. The 10-year probability of stroke for the baseline profile was similar between males and females, increasing from 0.6% for both sexes at age 40 to 2% for males and 2.5% for females are age 60. Among males, the estimated 10-year probability of stroke was higher for individuals with diabetes (age 40: 2.7%; age 60: 9.3%) than for current smokers (age 40: 1.1%; age 60: 3.8%) or those with elevated blood pressure (age 40: 1.4%; age 60: 5.1%). The highest risk was observed for the profile with high blood pressure and diabetes, with the estimated 10-year probability increasing from 6.7% at age 40 to 22.3% at age 60. Among females, the diabetes, elevated blood pressure, and current smoking profiles yielded similar estimated 10-year probabilities of stroke (age 40: 0.8%-1.1%; age 60: 3.6%-4.9%). The combined elevated blood pressure and diabetes profile again produced the highest estimated probabilities (age 40: 0.2%; age 60: 8.7%), although these estimates were notably lower than the corresponding values observed in males. It is important to note that the estimated probabilities were highly variable and formal tests for statistically significant differences between risk profiles were not performed.

```{r, include=TRUE}
table_risk_male
```

```{r, include=TRUE}
table_risk_female
```

```{r, include=TRUE}
table_time_varying
```




# Discussion and Conclusion

# Reproducible Research

## GitHub Link

`https://github.com/nageljo/BIOS6624`

## Dataset Filepath

`C:\Users\josep\OneDrive\AdvancedData\BIOS6624\Project3\Data\frmgham2.csv`


<!-- separate evaluation of males and females -->

<!-- backward selection of model covariates for males and females separately -->

<!-- descriptive investigation of how covariates change with time () -->

<!-- estimate risk (probability) of stroke for specific profiles -->

<!-- mention checking proportional hazards assumption -->



<!-- Do the analysis separately for males and female (the models can be different between sexes if there are different relationships) -->

<!-- Risk factors (baseline only) -->
<!--     include: age, diabetes, blood pressure (systolic) -->
<!--     several risk factors not sure about: coronary heart disease, blood pressure medication, smoking status (current Y/N), total cholesterol, BMI -->

<!-- Exclusion criteria: -->
<!--     only those who have not already had a stroke -->
<!--     check for outliers -->

<!-- only interested in strokes -->

<!-- We can treat events other than stroke as censored -->

<!-- we need descriptives of how predictors vary over time (we are not doing a time varying analysis) -->

<!-- Only check time varying for diabetes and blood pressure (use descriptives and interpret how a time varying analysis may change results) -->

<!-- censor events outside of the first 10 years of followup -->

<!-- We are mainly interested in estimate 10 year probability of stroke across risk profiles -->
<!--     Ages: 40, 50, 60 -->
<!--     Predict risk for average of continuous variables and 0 for binary separately for males and female -->

<!--     High blood pressure (160 systolic) -->
<!--     diabetes -->
<!--     high blood pressure + diabetes -->

<!--     you decided a risk profile (maybe obese) -->


<!-- do our covariates vary over time -->



<!-- # Reproducible Research -->

<!-- ## GitHub Link -->

<!-- `https://github.com/nageljo/BIOS6624` -->

<!-- ## Dataset Filepath -->

<!-- `C:\Users\josep\OneDrive\AdvancedData\BIOS6624\Project3\Data\frmgham2.csv` -->