---
title: "Sex-Specific Risk Factor Analysis and Probability Estimation for Stroke"
author: "Joe Nagel"
date: ""
output: 
  pdf_document:
    number_sections: true
fontsize: 12pt
mainfont: Helvetica
geometry: margin=1in
header-includes:
  - \usepackage{float}
  - \usepackage{setspace}
  - \doublespacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(list(echo = FALSE, include = FALSE, fig.width = 7.5))
```

```{r}
# read in packages
library(tidyverse)
library(patchwork)
library(naniar)
library(janitor)
library(knitr)
library(kableExtra)
library(gtsummary)
library(survival)
library(survminer)
library(broom)
TEN_YEARS = 365.25*10
```

```{r}
# read in data and create variables
(
    read_csv(
        '../Data/frmgham2.csv',
        show_col_types = FALSE,
        name_repair = "unique_quiet"
    )
    %>% clean_names()
    %>% mutate(
        .before = randid,
        stroke_in_followup = if_else(0 < timestrk & timestrk <= TEN_YEARS, stroke, 0),
        stroke_time = pmin(timestrk, TEN_YEARS)/365.25,
        sex = factor(sex, levels = 1:2, labels = c("Male", "Female")),
        bmi_cat = cut(
            bmi,
            breaks = c(-Inf, 25, 30, Inf),
            labels = c("Normal", "Overweight", "Obese")
        ),
        chol_cat = cut(
            totchol,
            breaks = c(-Inf, 240, Inf),
            labels = c("Normal", "High")
        )
    )
) -> fram
```

```{r}
# get total number of patients
fram$randid %>% unique() %>% length()
```

```{r}
# check for missingness
# naniar::vis_miss(fram)
```

```{r}
# create analysis dataset
fram_ana <- (
    fram
    # get baseline covariates and remove subjects who had a stroke before baseline
    %>% filter(period == 1 & prevstrk == 0)
    %>% select(randid, sex, stroke_in_followup, stroke_time, age, sysbp, totchol,
               bmi, diabetes, prevchd, bpmeds, cursmoke, bmi_cat, chol_cat)
)
```

```{r}
# create table 1
table_one <- (
    fram_ana
    %>% mutate(stroke_time = stroke_time)
    %>% tbl_summary(
        by = sex,
        statistic = list(
            all_continuous() ~ "{mean} ({sd})",
            all_categorical() ~ "{n} ({p}%)"
        ),
        missing = "ifany",
        missing_text = "Missing, n (%)",
        missing_stat = "{N_miss} ({p_miss}%)",
        include = c(
            stroke_in_followup,
            stroke_time,
            age,
            sysbp,
            totchol,
            bmi,
            diabetes,
            prevchd,
            bpmeds,
            cursmoke
        ),
        label = list(
            stroke_in_followup ~ "Stroke During Follow-up, n (%)",
            stroke_time ~ "Follow-up Time (Years), mean (SD)",
            age ~ "Age (Years), mean (SD)",
            sysbp ~ "Systolic Blood Pressure (mmHg), mean (SD)",
            totchol ~ "Total Cholesterol (mg/dL), mean (SD)",
            bmi ~ "BMI (kg/m²), mean (SD)",
            diabetes ~ "Diabetes, n (%)",
            prevchd ~ "Coronary Heart Disease, n (%)",
            bpmeds ~ "Blood Pressure Medication, n (%)",
            cursmoke ~ "Current Smoker, n (%)"
        ),
    )
    %>% add_overall(last = TRUE)
    %>% modify_footnote(everything() ~ NA)
    %>% as_kable_extra(
        latex = TRUE,
        booktabs = TRUE,
        linesep = "",
        caption = "Descriptive Statistics of Events, Follow-up, and Baseline Risk Factors",
        escape = FALSE
    )
    %>% kable_styling(latex_options = c("scale_down", "HOLD_position"))
)
```

```{r}
# plot KM curves for the five candidate risk variables
plot_km_curves <- (
    fram_ana
    %>% mutate(
        .keep = "none",
        stroke_time,
        stroke_in_followup,
        sex,
        `BP Meds` = factor(bpmeds, levels = 0:1, labels = c("No", "Yes")),
        `Current Smoker` = factor(cursmoke, levels = 0:1, labels = c("No", "Yes")),
        `Prev CHD` = factor(prevchd, levels = 0:1, labels = c("No", "Yes")),
        Cholesterol = chol_cat,
        BMI = bmi_cat
    )
    %>% drop_na()
    %>% pivot_longer(
        cols = !c(stroke_time, stroke_in_followup, sex),
        names_to = "variable",
        values_to = "value"
    )
    %>% group_by(sex, variable, value)
    %>% group_modify(~{
        fit <- survfit(Surv(stroke_time, stroke_in_followup) ~ 1, data = .x)
        tibble(time = fit$time, surv = fit$surv)
    })
    %>% ungroup()
    %>% ggplot(., aes(x = time, y = surv, color = value))
        + geom_step(linewidth = 0.9)
        + facet_grid(sex ~ variable)
        + labs(
            x = "Time (Years)",
            y = "Survival probability",
            color = "Group"
        )
        + scale_x_continuous(breaks = c(0, 5, 10))
        + theme_minimal()
)
```

```{r}
# fit full Cox PH models
full_formula = Surv(stroke_time, stroke_in_followup) ~ age + diabetes + sysbp + totchol + bmi + prevchd + bpmeds + cursmoke
base_formula = Surv(stroke_time, stroke_in_followup) ~ age + diabetes + sysbp

coxph_full_male <- coxph(
    full_formula,
    data = fram_ana %>% filter(sex == "Male") %>% drop_na()
)

coxph_full_female <- coxph(
    full_formula,
    data = fram_ana %>% filter(sex == "Female") %>% drop_na()
)
```

```{r}
# run backwards selection for males
step(
  coxph_full_male,
  scope = list(
    lower = base_formula,
    upper = full_formula
  ),
  direction = "backward"
)
```

```{r}
# backwards selection for females
step(
  coxph_full_female,
  scope = list(
    lower = base_formula,
    upper = full_formula
  ),
  direction = "backward"
)
```

```{r}
# fit final coxph models for males and females
coxph_final_male <- coxph(
    update(base_formula, . ~ . + cursmoke),
    data = fram_ana %>% filter(sex == "Male") %>% drop_na()
)

coxph_final_female <- coxph(
    update(base_formula, . ~ . + cursmoke),
    data = fram_ana %>% filter(sex == "Female") %>% drop_na()
)
```

```{r}
# create table of hazard ratios for the male and females models
table_model_coef <- (
    imap(
        list(male = coxph_final_male, female = coxph_final_female),
        ~ (
          tidy(.x, conf.int = TRUE, exponentiate = TRUE)
          %>% mutate(sex = .y, .before = 1)
        )
    )
    %>% list_rbind()
    %>% mutate(
        .keep = "none",
        term,
        sex,
        hazard_ratio = sprintf("%.3f", estimate),
        ci = sprintf("(%.3f, %.3f)", conf.low, conf.high),
        p_value = format.pval(p.value, digits = 2, eps = 0.001)
    )
    %>% pivot_wider(
        names_from = sex,
        values_from = c(hazard_ratio, ci, p_value)
    )
    %>% mutate(
        .keep = "unused",
        term = recode_values(
            term,
            "age" ~ "Age",
            "diabetes" ~ "Diabetes",
            "sysbp" ~ "Systolic Blood Pressure",
            "cursmoke" ~ "Smoking Status"
        )
    )
    %>% relocate(term,
                 hazard_ratio_male, ci_male, p_value_male,
                 hazard_ratio_female, ci_female, p_value_female)
    %>% kable(
        format = "latex",
        booktabs = TRUE,
        linesep  = "",
        caption  = "Cox Proportional Hazards Model Estimates",
        col.names = c("Coefficient",
                      "Hazard Ratio", "95% CI", "p-value",
                      "Hazard Ratio", "95% CI", "p-value"),
        align = "ccccc"
    )
    %>% row_spec(0, bold = TRUE)
    %>% add_header_above(c(" " = 1, "Male" = 3, "Female" = 3), bold = TRUE)
    %>% kable_styling(latex_options = c("scale_down", "HOLD_position"))
)
```

```{r}
# get Schoenfled residuals
coxph_res_male <- cox.zph(coxph_final_male)
coxph_res_female <- cox.zph(coxph_final_female)
```

```{r}
# check statistical test of Schoenfeld residuals for males
coxph_res_male$table
```

```{r}
# check statistical test of Schoenfeld residuals for females
coxph_res_female$table
```

```{r}
titles = c("Age (Years)", "Diabetes Status", "Systolic BP", "Smoking Status")

# plot Schoenfled residuals for males
(
    ggcoxzph(
        coxph_res_male,
        se = FALSE
    )
    %>% map2(
        titles,
        \(plot, title) {(
            plot
            + ylab("Schoenfeld Residuals")
            + ggtitle(title)
        )}
    )
    %>% wrap_plots(res_plots_male, ncol = 2)
        + plot_layout(axis_titles = "collect")
)
```

```{r}
# plot Schoenfeld residuals for females
(
    ggcoxzph(
        coxph_res_female,
        se = FALSE
    )
    %>% map2(
        titles,
        \(plot, title) {(
            plot
            + ylab("Schoenfeld Residuals")
            + ggtitle(title)
        )}
    )
    %>% wrap_plots(res_plots_male, ncol = 2)
        + plot_layout(axis_titles = "collect")
)
```

```{r}
# get mean blood pressure values across the population
population_bp_means = (
    fram_ana
    %>% summarise(
        .by = sex,
        sysbp = mean(sysbp)
    )
)
```

```{r}
# function to compute 10-year risk given a coxph model
compute_year10_risk <- function(model, newdata) {
    fit <- survfit(model, newdata)
    est <- summary(fit, times = 10)
    
    year10_risk <- tibble(
        risk = 1 - drop(est$surv),
        lower_cl = 1 - drop(est$upper),
        upper_cl = 1 - drop(est$lower)
    )
    
    return(year10_risk)
}
```

```{r}
# create risk profiles
risk_profiles <- (
    tribble(
        ~sex,      ~profile,                ~diabetes, ~cursmoke,
        "Male",   "Baseline",               0,         0,
        "Male",   "High BP",                0,         0,
        "Male",   "Diabetes",               1,         0,
        "Male",   "High BP + Diabetes",     1,         0,
        "Male",   "Smoker",                 0,         1,
        "Female", "Baseline",               0,         0,
        "Female", "High BP",                0,         0,
        "Female", "Diabetes",               1,         0,
        "Female", "High BP + Diabetes",     1,         0,
        "Female", "Smoker",                 0,         1
    )
    %>% left_join(population_bp_means, by = "sex")
    %>% mutate(sysbp = if_else(str_detect(profile, "High BP"), 160, sysbp))
    %>% crossing(age = c(40, 50, 60))
    %>% group_by(sex)
    %>% group_modify(~{
        model <- if (.y$sex[[1]] == "Male") coxph_final_male else coxph_final_female
        bind_cols(.x, compute_year10_risk(model, newdata = .x))
    })
    %>% ungroup()
)
```

```{r}
# create table of risk probabilities for males
table_risk_male <- (
    risk_profiles
    %>% filter(sex == "Male")
    %>% mutate(
        .keep = "none",
        `Risk Profile` = profile,
        age,
        risk_ci = sprintf("%03.1f (%03.1f, %03.1f)", 100*risk, 100*lower_cl, 100*upper_cl)
    )
    %>% pivot_wider(names_from = age, values_from = risk_ci)
    %>% kable(
        format = "latex",
        booktabs = TRUE,
        linesep  = "",
        caption  = "10-Year Probability of Stroke for Males",
        align = "cccc"
    )
    %>% row_spec(0, bold = TRUE)
    %>% add_header_above(c(" " = 1, "Age (Years)" = 3), bold = TRUE)
    %>% kable_styling(latex_options = c("scale_down", "HOLD_position"))
    %>% footnote(general = "Probabilities reported as % (95% CI)")
)
```

```{r}
# create table of risk probabilities for females
table_risk_female <- (
    risk_profiles
    %>% filter(sex == "Female")
    %>% mutate(
        .keep = "none",
        `Risk Profile` = profile,
        age,
        risk_ci = sprintf("%03.1f (%03.1f, %03.1f)", 100*risk, 100*lower_cl, 100*upper_cl)
    )
    %>% pivot_wider(names_from = age, values_from = risk_ci)
    %>% kable(
        format = "latex",
        booktabs = TRUE,
        linesep  = "",
        caption  = "10-Year Probability of Stroke for Females",
        align = "cccc"
    )
    %>% row_spec(0, bold = TRUE)
    %>% add_header_above(c(" " = 1, "Age (Years)" = 3), bold = TRUE)
    %>% kable_styling(latex_options = c("scale_down", "HOLD_position"))
    %>% footnote(general = "Probabilities reported as % (95% CI)")
)
```

```{r}
# create table of sysbp and diabetes over clinical exams
table_time_varying <- (
    fram
    %>% filter(prevstrk == 0)
    %>% summarise(
        .by = c(period, sex),
        mean_sysbp = mean(sysbp),
        sd_sysbp = sd(sysbp),
        n_diabetes = sum(diabetes),
        percent_diabetes = mean(diabetes)*100,
        n = n()
    )
    %>% mutate(
        .keep = "none",
        period,
        Sex = sex,
        `Systolic Blood Pressure (mmHg), mean (SD)` = sprintf("%.1f (%.1f)", mean_sysbp, sd_sysbp),
        `Diabetes, n (%)` = sprintf("%i (%.1f)", n_diabetes, percent_diabetes),
        `Number of Observations` = as.character(n)
    )
    %>% pivot_longer(cols = !c(period, Sex), names_to = "Characteristic")
    %>% mutate(
        Characteristic = factor(
            Characteristic,
            levels = c(
                "Number of Observations",
                "Systolic Blood Pressure (mmHg), mean (SD)",
                "Diabetes, n (%)")
        )
    )
    %>% relocate(Characteristic, Sex)
    %>% arrange(Characteristic, Sex, period)
    %>% pivot_wider(names_from = period, values_from = value)
    %>% kable(
        format = "latex",
        booktabs = TRUE,
        linesep  = "",
        caption  = "Summary of Blood Pressure and Diabetes Status by Clinical Exam",
        align = "ccccc"
    )
    %>% row_spec(0, bold = TRUE)
    %>% add_header_above(header = c(" " = 2, "Clinical Exam" = 3), bold = TRUE)
    %>% kable_styling(latex_options = c("scale_down", "HOLD_position"))
    %>% collapse_rows(columns = 1)
)
```

# Introduction

Strokes are medical emergencies caused by restricted or blocked blood flow to the brain. They are often caused by blood clots or tears in the blood vessels that supply blood to the brain. This analysis investigates the factors that contribute to the 10-year risk of stroke in adults. The data come from the Framingham Heart Study, a prospective study of 5,209 subjects living in Framingham, Massachusetts which began in 1948. The subset used in this analysis includes demographic information as well as risk factors and biological markers of heart disease which were collected during 3 clinical exams approximately 6 years apart. Each subject was also followed for 24 years for medical events related to heart disease: Angina Pectoris, Myocardial Infarction, Atherothrombotic Infarction, Cerebral Hemorrhage (Stroke), and death. The primary goal of this analysis is to identify risk factors of stroke and characterize the 10-year probability of stroke across specific risk profiles. Specifically, we will investigate the relationship between stroke and several potential risk factors: coronary heart disease (CHD), blood pressure medication, smoking status, total cholesterol, and body mass index (BMI). We will then model the 10-year probability of stroke separately for males and females as a function of baseline age, diabetes status, blood pressure and any additional risk factors we have identified. We will additionally investigate how the risk factors vary over time and any implications that variability might have for the results of the analysis.

# Methods

The analysis population included all participants in the Framingham cohort with no history of stroke at baseline; individuals with a prior stroke were excluded. Follow-up time was truncated at 10 years, and any observation time beyond 10 years was treated as censored at 10 years. For participants who experienced non-fatal, non-stroke events within the first 10 years (e.g., angina pectoris or myocardial infarction), follow-up continued, and fatal events were treated as censoring events. The primary risk factors of interest in this analysis were CHD, blood pressure medication, smoking status, total cholesterol, and BMI. All analysis variables were evaluated for missingness and outliers. A table of descriptive statistics of the analysis dataset was created. Additionally, all analyses were stratified by sex.

Kaplan–Meier curves were created to descriptively assess stroke-free survival across the risk factors of interest. Next, sex-stratified Cox proportional hazards models were fit, each containing the full set of candidate risk factors and adjusting for age, diabetes status, and systolic blood pressure. All models only included covariates measured at the first clinical exam (baseline). An AIC-based backward selection procedure was applied, with age, diabetes status, and systolic blood pressure retained as forced covariates. To improve interpretability and facilitate comparison by sex, the final male and female models retained any risk factor that was statistically significant in either sex. The proportional hazards assumption for the final models was checked using a combination of visualization and statistical testing of the Schoenfeld Residuals.

Final models were used to investigate the probability of stroke for a set of five distinct risk profiles. The first profile established a baseline risk and was defined as an individual who is a non-smoker, does not have diabetes or CHD, and does not use BP medication, with total cholesterol, BMI, and blood pressure set at population averages. This baseline was then compared to profiles reflecting individuals with high blood pressure (SBP = 160), diabetes, both high blood pressure and diabetes, and current smokers. The 10-year probability of stroke was calculated for each profile at ages 40, 50 and 60. Finally, an exploratory analysis examined changes in risk factors across clinical examinations, focusing on variability in systolic blood pressure and diabetes status over time. This assessment was used to evaluate potential limitations of relying solely on baseline measurements and to inform whether a time-varying covariate modeling approach might be warranted.

All analyses were performed using a Type I error rate ($\alpha$) of 0.05. All models and figures were created with R version 4.5.1. ChatGPT version 5.2 was used to debug code, outline the report, and copy-edit the report.

# Results

Of the original 4434 subjects, 32 had a history of stroke and were excluded from the analysis. The remaining 4402 comprised the analysis dataset. *Table 1* summarizes the baseline covariates, events, and follow-up time for these subjects. Smoking was more prevalent among men than women, with 61% of men and 40% of women reporting current smoking at baseline. For the complete case analysis, 127 subjects with missing BMI, total cholesterol, or blood pressure medication use values were dropped, resulting in a final sample size of 4275 subjects. High values of systolic blood pressure (295 mmHg), total cholesterol (696 mg/dL), and BMI (56 kg/m^2^) were observed. However, these values were biologically plausible and were retained for analysis.

```{r, include=TRUE}
table_one
```

The marginal stroke-free survival for each candidate risk factor is visualized in *Figure 1*. Blood pressure medication use and prevalent CHD appear to have the strongest relationship with stroke for both men and women. However, very few subjects used blood pressure medication (3.1%) or had CHD diagnosed before the study (4.2%). Notably, the relationship between high cholesterol (\> 240 mg/dL) and stroke appeared to be stronger among women than men.

```{r, include=TRUE, fig.cap="Kaplan-Meier Curves by Sex for Candidate Risk Factors"}
plot_km_curves
```

After AIC-based backward selection, the male model retained smoking status in addition to the fixed covariates, whereas the female model retained only the fixed covariates. For interpretability and comparability across sexes, the final reporting models for both males and females included age, diabetes status, systolic blood pressure, and smoking status. Both models were assessed for violations of the proportional hazards assumption. Visual inspection of the plots of the Schoenfleld residuals did not show signs of any violations and statistical tests of the residuals for the male (p = 0.345) and female (p = 0.979) models did not provide significant evidence of a violation. Estimated hazard ratios are shown in *Table 2*. Among males, age (p \< 0.001), diabetes (p \< 0.001), systolic blood pressure (p \< 0.001), and smoking status (p = 0.038) were significantly associated with the hazard of stroke. For males and adjusting for age, diabetes, and systolic blood pressure, smoking was associated with a 92% (3%, 257%) higher hazard of stroke. Among females, age (p \< 0.001) and systolic blood pressure (p \< 0.001) were significantly associated with stroke hazard, while the data provided insufficient evidence of a significant association for diabetes (p = 0.254) or smoking status (p = 0.207). For females and adjusting for age, diabetes, and systolic blood pressure, smoking was associated with a 44% (-18%, 252%) higher hazard of stroke.

```{r, include=TRUE}
table_model_coef
```

The final sex-specific models were used to estimate the 10-year probability of stroke for each of the five risk profiles. These estimates are presented for males in *Table 3* and for females in *Table 4*. The 10-year probability of stroke for the baseline profile was similar between males and females, increasing from 0.6% for both sexes at age 40 to 2% for males and 2.5% for females at age 60. Among males, the estimated 10-year probability of stroke was higher for individuals with diabetes (age 40: 2.7%; age 60: 9.3%) than for current smokers (age 40: 1.1%; age 60: 3.8%) or those with elevated blood pressure (age 40: 1.4%; age 60: 5.1%). The highest risk was observed for the profile with high blood pressure and diabetes, with the estimated 10-year probability increasing from 6.7% at age 40 to 22.3% at age 60. Among females, the diabetes, elevated blood pressure, and current smoking profiles yielded similar estimated 10-year probabilities of stroke (age 40: 0.8%-1.1%; age 60: 3.6%-4.9%). The combined elevated blood pressure and diabetes profile again produced the highest estimated probabilities (age 40: 0.2%; age 60: 8.7%), although these estimates were notably lower than the corresponding values observed in males. It is important to note that the estimated probabilities were highly variable and formal tests for significant differences between risk profiles were not performed.

```{r, include=TRUE}
table_risk_male
```

```{r, include=TRUE}
table_risk_female
```

Mean systolic blood pressure and the prevalence of diabetes across the three clinical examinations are summarized in Table 5. For both males and females, mean systolic blood pressure increased across successive exams. Among males, mean systolic blood pressure rose from 131.6 mmHg at Exam 1 to 138.9 mmHg at Exam 3. Females showed a similar increase, from 133.7 mmHg at Exam 1 to 140.6 mmHg at Exam 3. The prevalence of diabetes also increased over time in both groups. At Exam 1, diabetes was present in 3.1% of males and 2.4% of females; by Exam 3, these values increased to 8.3% in males and 7.1% in females.

```{r, include=TRUE}
table_time_varying
```

# Discussion and Conclusion

This analysis evaluated five candidate stroke risk factors: prevalent CHD, BP medication use, current smoking, total cholesterol, and BMI. Sex-stratified Cox proportional hazards models were fitted, adjusting for the fixed covariates of age, diabetes status, and systolic blood pressure. Following AIC-based backward selection, smoking status was the only candidate risk factor retained in the male model; the remaining candidate predictors did not materially improve model fit. In contrast, the female model retained only the fixed covariates, with none of the candidate risk factors improving model fit by AIC. However, these results should be interpreted cautiously because AIC-based variable selection can yield overly optimistic inference. Conventional inference assumes a pre-specified model and does not account for the selection step; as a result, post-selection p-values can be anti-conservative and uncertainty can be underestimated.

Using the final sex-specific models, 10-year stroke probabilities were calculated for five risk profiles. Across all ages and for both sexes, the baseline profile yielded the lowest estimated 10-year probability of stroke, whereas the combined high blood pressure and diabetes profile yielded the highest. Among males, the diabetes profile produced higher estimated probabilities than the high blood pressure and smoking profiles. Among females, the diabetes, high blood pressure, and smoking profiles had similar estimated probabilities.

An exploratory assessment of covariates across the three clinical examinations indicated that risk factor values did not remain static over time. Mean systolic blood pressure increased and diabetes prevalence also increased across clinical exams for both sexes. Because systolic blood pressure was significantly associated with stroke hazard in both males and females, these findings suggest that the analysis may be limited by relying only on baseline covariate measurements. Future work incorporating time-varying covariates may better capture the evolving risk profile over follow-up.

# Reproducible Research

## GitHub Link

`https://github.com/nageljo/BIOS6624`

## Dataset Filepath

`C:\Users\josep\OneDrive\AdvancedData\BIOS6624\Project3\Data\frmgham2.csv`
