---
title: "Risk Factors and Probability of Stroke"
author: "Joe Nagel"
date: ""
output: 
  pdf_document:
    number_sections: true
fontsize: 12pt
mainfont: Helvetica
geometry: margin=1in
header-includes:
  - \usepackage{float}
  - \usepackage{setspace}
  - \doublespacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(list(echo = FALSE, include = FALSE))
```

```{r}
# read in packages
library(tidyverse)
library(naniar)
library(janitor)
library(knitr)
library(kableExtra)
library(gtsummary)
library(survival)
library(survminer)
library(glmnet)
TEN_YEARS = 365.25*10
```

```{r}
(
    read_csv(
        '../Data/frmgham2.csv',
        show_col_types = FALSE,
        name_repair = "unique_quiet"
    )
    %>% clean_names()
    %>% mutate(
        .before = randid,
        stroke_in_followup = if_else(0 < timestrk & timestrk <= TEN_YEARS, stroke, 0),
        stroke_time = pmin(timestrk, TEN_YEARS)/365.25,
        sex = factor(sex, levels = 1:2, labels = c("Male", "Female")),
        bmi_cat = cut(
            bmi,
            breaks = c(-Inf, 25, 30, Inf),
            labels = c("Normal", "Overweight", "Obese")
        ),
        chol_cat = cut(
            totchol,
            breaks = c(-Inf, 240, Inf),
            labels = c("Normal", "High")
        )
    )
) -> fram
```

```{r}
# check for missingness
# naniar::vis_miss(fram)
```

```{r}
# create table 1
table_one <- (
    fram
    %>% filter(period == 1)
    %>% mutate(stroke_time = stroke_time)
    %>% tbl_summary(
        by = sex,
        statistic = list(
            all_continuous() ~ "{mean} ({sd})",
            all_categorical() ~ "{n} ({p}%)"
        ),
        missing = "ifany",
        missing_text = "Missing, n (%)",
        missing_stat = "{N_miss} ({p_miss}%)",
        include = c(
            stroke_in_followup,
            stroke_time,
            prevstrk,
            age,
            sysbp,
            totchol,
            bmi,
            diabetes,
            prevchd,
            bpmeds,
            cursmoke
        ),
        label = list(
            stroke_in_followup ~ "Stroke During Follow-up, n (%)",
            stroke_time ~ "Follow-up Time (Years), mean (SD)",
            prevstrk ~ "Prevalent Stroke, n (%)",
            age ~ "Age, mean (SD)",
            sysbp ~ "Systolic Blood Pressure, mean (SD)",
            totchol ~ "Total Cholesterol, mean (SD)",
            bmi ~ "BMI, mean (SD)",
            diabetes ~ "Diabetes, n (%)",
            prevchd ~ "Coronary Heart Disease, n (%)",
            bpmeds ~ "Blood Pressure Mediation, n (%)",
            cursmoke ~ "Current Smoker, n (%)"
        ),
    )
    %>% add_overall(last = TRUE)
    %>% modify_footnote(everything() ~ NA)
    %>% as_kable_extra(
        booktabs = TRUE,
        linesep = "",
        caption = "Descriptive Statistics of Events, Follow-up, and Baseline Risk Factors",
        escape = FALSE
    )
    %>% kable_styling(latex_options = c("scale_down", "HOLD_position"))
)
```


```{r}
# create analysis dataset
fram_ana <- (
    fram
    # get baseline covariates and remove subjects who had a stroke before baseline
    %>% filter(period == 1 & prevstrk == 0)
)
```

```{r}
get_km_plots <- function(variable, data) {
    surv_fit <- survfit(
        Surv(stroke_time, stroke_in_followup) ~ {{ variable }},
        data = data
    )
    
    min <- min(surv_fit$surv, na.rm = TRUE)
    
    curve_obj <- ggsurvplot(
        surv_fit,
        data = data,
        ylim = c(min - 0.05, 1)
    )
    
    plot <- curve_obj$plot
    
    return(plot)
}
```

```{r}
# get_km_plots(cursmoke, fram_ana)
```


```{r}
# c(chol_cat, bmi_cat, cursmoke, prevchd, bpmeds)
surv_fit <- survfit(Surv(stroke_time, stroke_in_followup) ~  bmi_cat, data = fram_ana)
```

```{r}
ggsurvplot(
    surv_fit,
    data = fram_ana,
    ylim = c(0.95, 1)
)$plot
```

```{r}
# fit mod

```


# Introduction

# Preliminary Methods

separate evaluation of males and females

backward selection of model covariates for males and females separately

descriptive investigation of how covariates change with time ()

estimate risk (probability) of stroke for specific profiles

mention checking proportional hazards assumption

```{r, include=TRUE}
table_one
```


<!-- # Results -->

<!-- # Discussion and Conclusion -->




Do the analysis separately for males and female (the models can be different between sexes if there are different relationships)

Risk factors (baseline only)
    include: age, diabetes, blood pressure (systolic)
    several risk factors not sure about: coronary heart disease, blood pressure medication, smoking status (current Y/N), total cholesterol, BMI

Exclusion criteria:
    only those who have not already had a stroke
    check for outliers

only interested in strokes

We can treat events other than stroke as censored

we need descriptives of how predictors vary over time (we are not doing a time varying analysis)

Only check time varying for diabetes and blood pressure (use descriptives and interpret how a time varying analysis may change results)

censor events outside of the first 10 years of followup

We are mainly interested in estimate 10 year probability of stroke across risk profiles
    Ages: 40, 50, 60
    Predict risk for average of continuous variables and 0 for binary separately for males and female
    
    High blood pressure (160 systolic)
    diabetes
    high blood pressure + diabetes
    
    you decided a risk profile (maybe obese)
    

do our covariates vary over time



# Reproducible Research

## GitHub Link

`https://github.com/nageljo/BIOS6624`

## Dataset Filepath

`C:\Users\josep\OneDrive\AdvancedData\BIOS6624\Project3\Data\frmgham2.csv`