---
title: "Project1"
author: "Joe Nagel"
date: ""
output: pdf_document
fontsize: 12pt
mainfont: Helvetica
geometry: margin=1in
header-includes:
  - \usepackage{float}
  - \usepackage{setspace}
  - \doublespacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(list(echo = FALSE, include = FALSE))
```

```{r}
# read in packages
source("functions.R")
library(tidyverse)
library(naniar)
library(janitor)
library(lme4)
library(lmerTest)
library(emmeans)
library(gtsummary)
library(gt)
library(knitr)
library(kableExtra)
library(patchwork) 
library(broom.mixed)
library(brms)
library(loo)
library(posterior)
library(bayestestR)
```

```{r}
(
    read_csv('../Data/hiv_6624_final.csv', show_col_types = FALSE, name_repair = "unique_quiet")
    %>% clean_names()
) -> hiv
```


```{r}
# read in data
(
    hiv
    # remove bad bmi values
    %>% mutate(
        log_vload = log10(vload),
        agg_ment = if_else(agg_ment < 0, NA_real_, agg_ment),
        bmi = if_else(bmi < 0 | 400 <= bmi, NA_real_, bmi),
        timepoint = if_else(years == 0, 'baseline', as.character(years)),
        adh = recode_values(
            adh,
            1:2 ~ ">95%",
            3:4 ~ "<95%",
            NA ~ NA_character_,
            default = "Not Handled"
        ) %>% factor(levels = c(">95%", "<95%")),
        hard_drugs = recode_values(
            hard_drugs,
            0 ~ "No Hard Drug Use",
            1 ~ "Hard Drug Use",
            NA ~ NA_character_,
            default = "Not Handled"
        ) %>% factor(levels = c("No Hard Drug Use", "Hard Drug Use")),
        smoke = recode_values(
            smoke,
            1:2 ~ "Never/Former Smoker",
            3 ~ "Current Smoker",
            NA ~ NA_character_,
            default = "Not Handled"
        ) %>% factor(levels = c("Never/Former Smoker", "Current Smoker")),
        race_recode = recode_values(
            race,
            1 ~ "Non-Hispanic White",
            2:8 ~ "Other",
            NA ~ NA_character_,
            default = "Not Handled"
        ) %>% factor(levels = c("Non-Hispanic White", "Other")),
        edu_recode = recode_values(
            educbas,
            1:4 ~ "No College Degree",
            5:7 ~ "College Degree or Greater",
            NA ~ NA_character_,
            default = "Not Handled"
        ) %>% factor(levels = c("No College Degree", "College Degree or Greater"))
    )
    
    # hiv positive only
    %>% filter(hivpos == 1)
    
    # filter to baseline and year 2
    %>% filter(years %in% c(0, 2))
    %>% select(
        newid, timepoint, agg_ment, agg_phys, leu3n, log_vload,
        hard_drugs, adh, age, bmi, smoke, edu_recode, race_recode
    )
    %>% pivot_wider(
        id_cols = newid,
        names_from = timepoint,
        names_glue = "{.value}_{timepoint}",
        values_from = !c(newid, timepoint)
    )
    
    # calculate differences between baseline and year 2
    %>% mutate(
        # .keep = 'unused',
        .after = newid,
        diff_agg_ment = agg_ment_2 - agg_ment_baseline,
        diff_agg_phys = agg_phys_2 - agg_phys_baseline,
        diff_leu3n = leu3n_2 - leu3n_baseline,
        diff_log_vload = log_vload_2 - log_vload_baseline,
        adh = adh_2
    )
    %>% select(-matches('_2'), -adh_baseline)
    
    # keep only participants with all outcomes at 0 and 2 years
    %>% filter(!if_any(matches('diff'), is.na))
) -> hiv_clean
```

```{r}
(
    hiv_clean
    %>% mutate(
        missing = if_any(matches("diff"), is.na)
    )
    %>% summarise(
        .by = c(hard_drugs_baseline, missing),
        n = n()
    )
    %>% group_by(hard_drugs_baseline)
    %>% mutate(
        prop = n/sum(n)
    )
    %>% arrange(hard_drugs_baseline, missing)
)
```


```{r}
# create table 1
table1 <- (
    hiv_clean
    %>% tbl_summary(
        by = hard_drugs_baseline,
        statistic = list(
            all_continuous() ~ "{mean} ({sd})",
            all_categorical() ~ "{n} ({p}%)"
        ),
        missing = "ifany",
        missing_text = "Missing/Improbable, n (%)",
        missing_stat = "{N_miss} ({p_miss}%)",
        label = list(                              
            agg_ment_baseline    ~ "SF-36 MCS, mean (SD)",
            agg_phys_baseline    ~ "SF-36 PCS, mean (SD)",
            leu3n_baseline       ~ "CD4+ T Cell Count, mean (SD)",
            log_vload_baseline   ~ "Log Viral Load, mean (SD)",
            adh                  ~ "Adherence, n (%)",
            age_baseline         ~ "Age (Years), mean (SD)",
            bmi_baseline         ~ "BMI, mean (SD)",
            smoke_baseline       ~ "Smoking Status, n (%)",
            edu_recode_baseline  ~ "Education, n (%)",
            race_recode_baseline ~ "Race/Ethnicity, n (%)"
        ),
        include = c(
            agg_ment_baseline,
            agg_phys_baseline,
            leu3n_baseline,
            log_vload_baseline,
            age_baseline,
            bmi_baseline,
            adh,
            smoke_baseline,
            edu_recode_baseline,
            race_recode_baseline
        ),
        digits = list(
            all_continuous() ~ 1
            # all_categorical() ~ 1
        )
    )
    %>% add_overall(last = TRUE)
    %>% modify_footnote(everything() ~ NA)
    %>% as_kable_extra(
        booktabs = TRUE,
        linesep = "",
        caption = "Descriptive Statistics of Baseline Covariates and Year 2 Adherence",
        escape = TRUE
    )
    %>% kable_styling(latex_options = c("scale_down", "HOLD_position"))
)
```

```{r, include = TRUE}
table1
```

```{r}
# model formulas
formula_ment <- diff_agg_ment ~
    hard_drugs_baseline + agg_ment_baseline + age_baseline +
    bmi_baseline + smoke_baseline + edu_recode_baseline + race_recode_baseline

formula_phys <- diff_agg_phys ~
    hard_drugs_baseline + agg_phys_baseline + age_baseline +
    bmi_baseline + smoke_baseline + edu_recode_baseline + race_recode_baseline

formula_vload <- diff_log_vload ~
    hard_drugs_baseline + log_vload_baseline + age_baseline +
    bmi_baseline + smoke_baseline + edu_recode_baseline + race_recode_baseline

formula_leu3n <- diff_leu3n ~
    hard_drugs_baseline + leu3n_baseline + age_baseline +
    bmi_baseline + smoke_baseline + edu_recode_baseline + race_recode_baseline
```

```{r}
# frequentist models
fmod_ment <- lm(formula_ment, data = hiv_clean)
fmod_phys <- lm(formula_phys, data = hiv_clean)
fmod_vload <- lm(formula_vload, data = hiv_clean)
fmod_leu3n <- lm(formula_leu3n, data = hiv_clean)
```

```{r}
# frequentist models with adherence
fmod_ment <- lm(update(formula_ment, . ~ . + adh), data = hiv_clean)
fmod_phys <- lm(update(formula_phys, . ~ . + adh), data = hiv_clean)
fmod_vload <- lm(update(formula_vload, . ~ . + adh), data = hiv_clean)
fmod_leu3n <- lm(update(formula_leu3n, . ~ . + adh), data = hiv_clean)
```

```{r}
# fit Bayesian models
bmod_ment <- fit_brm(formula = formula_ment, data = hiv_clean, file = "../Models/bmod_ment")
bmod_phys <- fit_brm(formula = formula_phys, data = hiv_clean, file = "../Models/bmod_phys")
bmod_vload <- fit_brm(formula = formula_vload, data = hiv_clean, file = "../Models/bmod_vload")
bmod_leu3n <- fit_brm(formula = formula_leu3n, data = hiv_clean, file = "../Models/bmod_leu3n")
```

```{r}
# fit Bayesian models without hard drug term
bmod_ment_red <- fit_brm(
    formula = update(formula_ment, . ~ . - hard_drugs_baseline),
    data = hiv_clean,
    file = "../Models/bmod_ment_red"
)
bmod_phys_red <- fit_brm(
    formula = update(formula_phys, . ~ . - hard_drugs_baseline),
    data = hiv_clean,
    file = "../Models/bmod_phys_red"
)
bmod_vload_red <- fit_brm(
    formula = update(formula_vload, . ~ . - hard_drugs_baseline),
    data = hiv_clean,
    file = "../Models/bmod_vload_red"
)
bmod_leu3n_red <- fit_brm(
    formula = update(formula_leu3n, . ~ . - hard_drugs_baseline),
    data = hiv_clean,
    file = "../Models/bmod_leu3n_red"
)
```

```{r}
# fit Bayesian models with adherence term
bmod_ment_adh <- fit_brm(
    formula = update(formula_ment, . ~ . + adh),
    data = hiv_clean,
    file = "../Models/bmod_ment_adh"
)
bmod_phys_adh <- fit_brm(
    formula = update(formula_phys, . ~ . + adh),
    data = hiv_clean,
    file = "../Models/bmod_phys_adh"
)
bmod_vload_adh <- fit_brm(
    formula = update(formula_vload, . ~ . + adh),
    data = hiv_clean,
    file = "../Models/bmod_vload_adh"
)
bmod_leu3n_adh <- fit_brm(
    formula = update(formula_leu3n, . ~ . + adh),
    data = hiv_clean,
    file = "../Models/bmod_leu3n_adh"
)
```

```{r}
# Named list to iterate across outcomes (for labeling rows)
row_labels <- c(
  "SF-36 MCS"   = "diff_agg_ment",
  "SF-36 PCS" = "diff_agg_phys",
  "CD4+ T Cell Count"  = "diff_leu3n",
  "Log Viral Load"     = "diff_log_vload"
)

# Collect models in parallel lists (order must match row_labels)
fmods      <- list(fmod_ment, fmod_phys, fmod_leu3n, fmod_vload)  
bmods_full <- list(bmod_ment, bmod_phys, bmod_leu3n, bmod_vload)
bmods_red  <- list(bmod_ment_red, bmod_phys_red, bmod_leu3n_red, bmod_vload_red)

table_freq_bayes <- (
  pmap_dfr(
    list(names(row_labels), fmods, bmods_full, bmods_red),
    function(outcome_label, ffit, bfit_full, bfit_red) {
      # Frequentist
      fr    <- extract_lm_stats(ffit, "hard_drugs_baseline")
      fr_ci <- sprintf("[%s, %s]", fmt(fr$lcl), fmt(fr$ucl))
      p_str <- if (fr$p < 0.001) "<0.001" else fmt(fr$p, digits = 3)

      # Bayesian with HDI (from full model)
      br    <- extract_brm_stats_hdi(bfit_full, "hard_drugs_baseline", cred = 0.95)
      br_ci <- sprintf("[%s, %s]", fmt(br$lcl), fmt(br$ucl))

      # ΔLOOIC using prefit reduced model
      dL <- delta_loo_prefit(bfit_full, bfit_red, reloo = FALSE)

      tibble::tibble(
        Outcome                  = outcome_label,
        `Frequentist Estimate`   = as.numeric(fr$estimate),
        `Frequentist 95% CI`     = fr_ci,
        `p-value`                = p_str,
        `Bayesian Estimate`      = as.numeric(br$estimate),
        `Bayesian 95% HDI`       = br_ci,
        `delta LOOIC`            = as.numeric(dL$delta_looic)
      )
    }
  ) %>%
  mutate(
    `Frequentist Estimate` = as.numeric(fmt(`Frequentist Estimate`)),
    `Bayesian Estimate`    = as.numeric(fmt(`Bayesian Estimate`)),
    `delta LOOIC`          = as.numeric(fmt(`delta LOOIC`, digits = 2))
  ) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    linesep  = "",
    caption  = "Frequentist and Bayesian Model Results for Hard Drug Use Effect",
    align    = "lcccccc",
    escape   = FALSE,
    col.names = c("Outcome", "Estimate", "95\\% CI", "p-value",
                  "Estimate", "95\\% HDI", "$\\Delta$ LOO-IC")
  ) %>%
  add_header_above(
    c(" " = 1, "Frequentist" = 3, "Bayesian" = 3),
    bold = TRUE
  ) %>%
  kable_styling(latex_options = c("scale_down", "HOLD_position"))
)
```


```{r, include=TRUE}
table_freq_bayes
```


```{r}
summary(fmod_leu3n)
```


```{r}
hiv_clean %>%
  # Pivot the difference columns longer so we can plot them all at once
  pivot_longer(cols = starts_with("diff"), names_to = "outcome", values_to = "change") %>%
  ggplot(aes(x = hard_drugs_baseline, y = change, fill = hard_drugs_baseline)) +
  geom_boxplot(na.rm = TRUE) +
  facet_wrap(~outcome, scales = "free_y") +
  theme_minimal() +
  labs(title = "Treatment Response (Year 0 to 2) by Baseline Hard Drug Use",
       y = "Change from Baseline")
```


# Introduction

HIV (human immunodeficiency virus) is a virus that attacks the immune system; it destroys T-cells and inhibits your body's ability to fight infection. This analysis evaluates the effectiveness of highly active antiretroviral treatment (HAART), the standard treatment for HIV patients. The data come from the Multicenter AIDS Cohort Study, an ongoing study investigating HIV infection in homosexual and bisexual men in the United States. The dataset includes longitudinal demographic information, laboratory measurements, and quality of life measurements from the 36-Item Short Form Health Survey (SF-36) on men infected with HIV. Data was collected on subjects at the beginning of HAART treatment (baseline) and every year thereafter for up to 8 years. The primary goal of this analysis is to characterize how treatment response after 2 years of treatment differs based on a subjects hard drug use status at baseline. Specifically, we will investigate the difference in SF-36 Mental Component Summary (MCS) score, SF-36 Physical Component Summary (PCS) score, HIV viral load, and CD4+ T cell count between baseline and two years across hard drug use status at baseline while adjusting for baseline outcomes, age, bmi, smoking status, education, and race. Additionally, we will investigate if any significant differences across drug use are explained by study protocol adherence. 

# Preliminary Methods

We will begin by checking the analysis variables for missingness and outliers; we will specifically check if missingness differs between hard drug users and non-hard drug uses. We will calculate the change in outcomes as the value at Year 2 minus the value at Year 0. We will also visualize the distributions of the four outcome difference scores using histograms and boxplots to identify potential outliers and assess the need for any transformations prior to analysis. For the primary analysis, we will generate descriptive statistics to compare baseline demographic and clinical characteristics between those who reported hard drug use and those who did not. To evaluate the effect of baseline hard drug use on treatment response, we will fit four separate multivariable linear regression models—one for each outcome (change in PCS, MCS, CD4+ count, and viral load). The primary predictor of interest will be baseline hard drug use status. All models will adjust for the specified demographic variables: age, BMI, smoking status, education, and race. We will approach modeling from both a Bayesian and non-Bayesian approach. For the non-Bayesian, we will use standard Ordinary Least Squares (OLS) regression, reporting point estimates, 95% confidence intervals, and p-values. For the Bayesian, we will fit Bayesian linear regression models. We will specify non-informative or weakly informative prior distributions for our parameters and estimate posterior distributions using Markov Chain Monte Carlo (MCMC) methods. We will report posterior means, 95% Highest Posterior Density (HPD) credible intervals, and posterior probabilities to evaluate the primary research question. We will compare the results from both frameworks and determine if there are any differences in the interpretations. Finally, we will investigate if any relationship between hard drug use and the outcomes is explained by adherence to the treatment protocol.

# Methods

This analysis only considered subjects with all outcomes (PCS, MCS, CD4+ count, and viral load) measured at both baseline and two years. We assessed if missingness in the outcomes differed between hard drug users and non hard drug users. Due to low sample sizes within groups, Race/Ethnicity was combined into Non-Hispanic White and Other and Education was combined into No College Degree and College Degree or Greater. Outlier BMI measurements ($999$ (insufficient data), $-1$ (improbable value), and any biologically improssible values (BMI $<0$ or $>400$)) were excluded from the analysis. Additionally, viral load was analysed on the $\log_{10}$ scale 


Potential Limitation of this work: it could be that hard drug users have better adherence because the ones that dont have good adherence did not have measurements at year 2