---
title: "Impacts of Hard Drug Use on HAART Efficacy in HIV-Positive Men"
author: "Joe Nagel"
date: ""
output: 
  pdf_document:
    number_sections: true
fontsize: 12pt
mainfont: Helvetica
geometry: margin=1in
header-includes:
  - \usepackage{float}
  - \usepackage{setspace}
  - \doublespacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(list(echo = FALSE, include = FALSE))
```

```{r}
# read in packages
source("functions.R")
library(tidyverse)
library(naniar)
library(janitor)
library(lme4)
library(lmerTest)
library(emmeans)
library(gtsummary)
library(gt)
library(knitr)
library(kableExtra)
library(patchwork) 
library(broom.mixed)
library(brms)
library(loo)
library(posterior)
library(bayestestR)
```

```{r}
(
    read_csv(
        '../Data/hiv_6624_final.csv',
        show_col_types = FALSE,
        name_repair = "unique_quiet"
    )
    %>% clean_names()
) -> hiv
```

```{r}
# read in data
(
    hiv
    # remove bad bmi values
    %>% mutate(
        log_vload = log10(vload),
        agg_ment = if_else(agg_ment < 0, NA_real_, agg_ment),
        bmi = if_else(bmi < 0 | 400 <= bmi, NA_real_, bmi),
        timepoint = if_else(years == 0, 'baseline', as.character(years)),
        adh = recode_values(
            adh,
            1:2 ~ ">=95%",
            3:4 ~ "<95%",
            NA ~ NA_character_,
            default = "Not Handled"
        ) %>% factor(levels = c(">=95%", "<95%")),
        hard_drugs = recode_values(
            hard_drugs,
            0 ~ "No Hard Drug Use",
            1 ~ "Hard Drug Use",
            NA ~ NA_character_,
            default = "Not Handled"
        ) %>% factor(levels = c("No Hard Drug Use", "Hard Drug Use")),
        smoke = recode_values(
            smoke,
            1:2 ~ "Never/Former Smoker",
            3 ~ "Current Smoker",
            NA ~ NA_character_,
            default = "Not Handled"
        ) %>% factor(levels = c("Never/Former Smoker", "Current Smoker")),
        race_recode = recode_values(
            race,
            1 ~ "Non-Hispanic White",
            2:8 ~ "Other",
            NA ~ NA_character_,
            default = "Not Handled"
        ) %>% factor(levels = c("Non-Hispanic White", "Other")),
        edu_recode = recode_values(
            educbas,
            1:4 ~ "No College Degree",
            5:7 ~ "College Degree or Greater",
            NA ~ NA_character_,
            default = "Not Handled"
        ) %>% factor(levels = c("No College Degree", "College Degree or Greater"))
    )
    
    # hiv positive only
    %>% filter(hivpos == 1)
    
    # filter to baseline and year 2
    %>% filter(years %in% c(0, 2))
    %>% select(
        newid, timepoint, agg_ment, agg_phys, leu3n, log_vload,
        hard_drugs, adh, age, bmi, smoke, edu_recode, race_recode
    )
    %>% pivot_wider(
        id_cols = newid,
        names_from = timepoint,
        names_glue = "{.value}_{timepoint}",
        values_from = !c(newid, timepoint)
    )
    
    # calculate differences between baseline and year 2
    %>% mutate(
        # .keep = 'unused',
        .after = newid,
        diff_agg_ment = agg_ment_2 - agg_ment_baseline,
        diff_agg_phys = agg_phys_2 - agg_phys_baseline,
        diff_leu3n = leu3n_2 - leu3n_baseline,
        diff_log_vload = log_vload_2 - log_vload_baseline,
        adh = adh_2
    )
    %>% select(-matches('_2'), -adh_baseline)
) -> hiv_clean
```

```{r}
hiv_analysis <- hiv_clean %>% filter(!if_any(matches('diff'), is.na))
```

```{r}
nrow(hiv_clean)
nrow(hiv_analysis)
```

```{r, eval = FALSE}
hist(hiv_analysis$diff_agg_ment)
hist(hiv_analysis$diff_agg_phys)
hist(hiv_analysis$diff_log_vload)
hist(hiv_analysis$diff_leu3n)
```

```{r}
# check for missingness in outcomes across hard drug use
(
    hiv_clean
    %>% summarise(
        .by = hard_drugs_baseline,
        n_missing = sum(if_any(matches("diff"), is.na)),
        total = n(),
        prop = mean(if_any(matches("diff"), is.na))
    )
)
```


```{r}
# create table 1
table1 <- (
    hiv_analysis
    %>% tbl_summary(
        by = hard_drugs_baseline,
        statistic = list(
            all_continuous() ~ "{mean} ({sd})",
            all_categorical() ~ "{n} ({p}%)"
        ),
        missing = "ifany",
        missing_text = "Missing/Improbable, n (%)",
        missing_stat = "{N_miss} ({p_miss}%)",
        label = list(                              
            agg_ment_baseline    ~ "SF-36 MCS, mean (SD)",
            agg_phys_baseline    ~ "SF-36 PCS, mean (SD)",
            leu3n_baseline       ~ "CD4+ T Cell Count, mean (SD)",
            log_vload_baseline   ~ "Log10 Viral Load, mean (SD)",
            adh                  ~ "Adherence, n (%)",
            age_baseline         ~ "Age (Years), mean (SD)",
            bmi_baseline         ~ "BMI, mean (SD)",
            smoke_baseline       ~ "Smoking Status, n (%)",
            edu_recode_baseline  ~ "Education, n (%)",
            race_recode_baseline ~ "Race/Ethnicity, n (%)"
        ),
        include = c(
            agg_ment_baseline,
            agg_phys_baseline,
            leu3n_baseline,
            log_vload_baseline,
            age_baseline,
            bmi_baseline,
            adh,
            smoke_baseline,
            edu_recode_baseline,
            race_recode_baseline
        ),
        digits = list(
            all_continuous() ~ 1
            # all_categorical() ~ 1
        )
    )
    %>% add_overall(last = TRUE)
    
    # ADDED THIS BLOCK TO SPLIT THE HEADERS
    %>% modify_header(
        all_stat_cols() ~ "{level}\nN = {n}",
        stat_0 ~ "Overall\nN = {N}"
    )
    
    %>% modify_footnote(everything() ~ NA)
    %>% as_kable_extra(
        booktabs = TRUE,
        linesep = "",
        caption = "Descriptive Statistics of Baseline Covariates and Year 2 Adherence",
        escape = FALSE
    )
    %>% kable_styling(latex_options = c("scale_down", "HOLD_position"))
)
```

```{r}
# model formulas
formula_ment <- diff_agg_ment ~
    hard_drugs_baseline + agg_ment_baseline + age_baseline +
    bmi_baseline + smoke_baseline + edu_recode_baseline + race_recode_baseline

formula_phys <- diff_agg_phys ~
    hard_drugs_baseline + agg_phys_baseline + age_baseline +
    bmi_baseline + smoke_baseline + edu_recode_baseline + race_recode_baseline

formula_vload <- diff_log_vload ~
    hard_drugs_baseline + log_vload_baseline + age_baseline +
    bmi_baseline + smoke_baseline + edu_recode_baseline + race_recode_baseline

formula_leu3n <- diff_leu3n ~
    hard_drugs_baseline + leu3n_baseline + age_baseline +
    bmi_baseline + smoke_baseline + edu_recode_baseline + race_recode_baseline
```

```{r}
# frequentist models
fmod_ment <- lm(formula_ment, data = hiv_analysis)
fmod_phys <- lm(formula_phys, data = hiv_analysis)
fmod_vload <- lm(formula_vload, data = hiv_analysis)
fmod_leu3n <- lm(formula_leu3n, data = hiv_analysis)
```

```{r}
# frequentist models with adherence
fmod_ment_adh <- lm(update(formula_ment, . ~ . + adh), data = hiv_analysis)
fmod_phys_adh <- lm(update(formula_phys, . ~ . + adh), data = hiv_analysis)
fmod_vload_adh <- lm(update(formula_vload, . ~ . + adh), data = hiv_analysis)
fmod_leu3n_adh <- lm(update(formula_leu3n, . ~ . + adh), data = hiv_analysis)
```

```{r}
# fit Bayesian models
bmod_ment <- fit_brm(formula = formula_ment, data = hiv_analysis, file = "../Models/bmod_ment")
bmod_phys <- fit_brm(formula = formula_phys, data = hiv_analysis, file = "../Models/bmod_phys")
bmod_vload <- fit_brm(formula = formula_vload, data = hiv_analysis, file = "../Models/bmod_vload")
bmod_leu3n <- fit_brm(formula = formula_leu3n, data = hiv_analysis, file = "../Models/bmod_leu3n")
```

```{r}
# fit Bayesian models without hard drug term
bmod_ment_red <- fit_brm(
    formula = update(formula_ment, . ~ . - hard_drugs_baseline),
    data = hiv_analysis,
    file = "../Models/bmod_ment_red"
)
bmod_phys_red <- fit_brm(
    formula = update(formula_phys, . ~ . - hard_drugs_baseline),
    data = hiv_analysis,
    file = "../Models/bmod_phys_red"
)
bmod_vload_red <- fit_brm(
    formula = update(formula_vload, . ~ . - hard_drugs_baseline),
    data = hiv_analysis,
    file = "../Models/bmod_vload_red"
)
bmod_leu3n_red <- fit_brm(
    formula = update(formula_leu3n, . ~ . - hard_drugs_baseline),
    data = hiv_analysis,
    file = "../Models/bmod_leu3n_red"
)
```

```{r}
# fit Bayesian models with adherence term
bmod_ment_adh <- fit_brm(
    formula = update(formula_ment, . ~ . + adh),
    data = hiv_analysis,
    file = "../Models/bmod_ment_adh"
)
bmod_phys_adh <- fit_brm(
    formula = update(formula_phys, . ~ . + adh),
    data = hiv_analysis,
    file = "../Models/bmod_phys_adh"
)
bmod_vload_adh <- fit_brm(
    formula = update(formula_vload, . ~ . + adh),
    data = hiv_analysis,
    file = "../Models/bmod_vload_adh"
)
bmod_leu3n_adh <- fit_brm(
    formula = update(formula_leu3n, . ~ . + adh),
    data = hiv_analysis,
    file = "../Models/bmod_leu3n_adh"
)
```

```{r}
# create table of hard drug use effects for Bayesian and frequentist models

# Named list to iterate across outcomes (for labeling rows)
row_labels <- c(
  "SF-36 MCS"   = "diff_agg_ment",
  "SF-36 PCS" = "diff_agg_phys",
  "CD4+ Count"  = "diff_leu3n",
  "Log Viral Load"     = "diff_log_vload"
)

# Clinical thresholds matching the order of row_labels
clinical_thresholds <- c(2, 2, 50, 0.5)

# Collect models in parallel lists (order must match row_labels)
fmods      <- list(fmod_ment, fmod_phys, fmod_leu3n, fmod_vload)  
bmods_full <- list(bmod_ment, bmod_phys, bmod_leu3n, bmod_vload)
bmods_red  <- list(bmod_ment_red, bmod_phys_red, bmod_leu3n_red, bmod_vload_red)

table_freq_bayes <- (
  pmap_dfr(
    list(names(row_labels), fmods, bmods_full, bmods_red, clinical_thresholds),
    function(outcome_label, ffit, bfit_full, bfit_red, thresh) {
      
      # Frequentist
      fr    <- extract_lm_stats(ffit, "hard_drugs_baseline")
      fr_ci <- sprintf("[%s, %s]", fmt(fr$lcl), fmt(fr$ucl))
      p_str <- if (fr$p < 0.001) "<0.001" else fmt(fr$p, digits = 3)

      # Bayesian with HDI and clinical threshold
      br    <- extract_brm_stats_hdi(bfit_full, "hard_drugs_baseline", cred = 0.95, threshold = thresh)
      br_ci <- sprintf("[%s, %s]", fmt(br$lcl), fmt(br$ucl))
      p_clin_str <- fmt(br$p_clin, digits = 3)

      # ΔLOOIC using prefit reduced model
      dL <- delta_loo_prefit(bfit_full, bfit_red, reloo = FALSE)

      tibble::tibble(
        Outcome                   = outcome_label,
        `Frequentist Estimate`    = as.numeric(fr$estimate),
        `Frequentist 95% CI`      = fr_ci,
        `p-value`                 = p_str,
        `Bayesian Estimate`       = as.numeric(br$estimate),
        `Bayesian 95% HDI`        = br_ci,
        `P(|Effect| > Threshold)` = p_clin_str,
        `delta LOOIC`             = as.numeric(dL$delta_looic)
      )
    }
  ) %>%
  mutate(
    `Frequentist Estimate` = as.numeric(fmt(`Frequentist Estimate`)),
    `Bayesian Estimate`    = as.numeric(fmt(`Bayesian Estimate`)),
    `delta LOOIC`          = as.numeric(fmt(`delta LOOIC`, digits = 2))
  ) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    linesep  = "",
    caption  = "Frequentist and Bayesian Estimates for Hard Drug Use Effects",
    align    = "lcccccccc", # Added one 'c' for the new column
    escape   = FALSE,
    col.names = c("Outcome", "Estimate", "95\\% CI", "p-value",
                  "Estimate", "95\\% HDI", "Posterior Probability", "$\\Delta$ LOO-IC")
  ) %>%
  add_header_above(
    c(" " = 1, "Frequentist" = 3, "Bayesian" = 4), # Bayesian span goes from 3 to 4
    bold = TRUE
  ) %>%
  kable_styling(latex_options = c("scale_down", "HOLD_position")) %>%
  footnote(general = "The posterior probability represents the probability of a clinically meaningful effect or greater.")
)
```

```{r}
# create Bayesian hard drug use and adherence table

# Collect models in parallel lists matching the outcomes
bmods_base <- list(bmod_ment, bmod_phys, bmod_leu3n, bmod_vload)
bmods_adh  <- list(bmod_ment_adh, bmod_phys_adh, bmod_leu3n_adh, bmod_vload_adh)

row_labels <- c(
  "SF-36 MCS",
  "SF-36 PCS",
  "CD4+ T Cell Count",
  "Log Viral Load"
)

table_adh_comparison_bayes <- pmap_dfr(
  list(row_labels, bmods_base, bmods_adh),
  function(outcome_label, fit_base, fit_adh) {
    
    # 1. Unadjusted Hard drugs effect (Base model without adherence)
    res_no_adh <- extract_brm_stats_hdi(fit_base, "hard_drugs_baseline")
    
    # 2. Adjusted Hard drugs effect (Model with adherence)
    res_adh <- extract_brm_stats_hdi(fit_adh, "hard_drugs_baseline")
    
    # 3. Adherence effect (Model with adherence)
    res_adh_term <- extract_brm_stats_hdi(fit_adh, "adh")
    
    # 4. Delta LOOIC
    dL <- delta_loo_prefit(full_fit = fit_adh, reduced_fit = fit_base, reloo = FALSE)
    
    tibble::tibble(
      Outcome = outcome_label,
      `Unadj_Est` = fmt(res_no_adh$estimate, 2),
      `Unadj_HDI` = sprintf("[%s, %s]", fmt(res_no_adh$lcl, 2), fmt(res_no_adh$ucl, 2)), 
      `Adj_Est` = fmt(res_adh$estimate, 2),
      `Adj_HDI` = sprintf("[%s, %s]", fmt(res_adh$lcl, 2), fmt(res_adh$ucl, 2)), 
      `adh_est` = fmt(res_adh_term$estimate, 2),
      `adh_HDI  ` = sprintf("[%s, %s]", fmt(res_adh_term$lcl, 2), fmt(res_adh_term$ucl, 2)),
      `delta_loo` = fmt(dL$delta_looic, 1)
    )
  }
) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    linesep = "",
    caption = "Bayesian Estimates: Impact of Adjusting for Adherence on the Effect of Baseline Hard Drug Use (HDU)",
    escape = FALSE,
    align = "lccccccc", # Updated to exactly match 8 columns
    col.names = c(
      "Outcome", 
      "Estimate", 
      "95\\% HDI", 
      "Estimate", 
      "95\\% HDI", 
      "Estimate", 
      "95\\% HDI", 
      "$\\Delta$ LOO-IC"
    )
  ) %>%
  add_header_above(
    c(
        " " = 1,
        "Unadjusted HDU Effect" = 2, 
        "Adjusted HDU Effect" = 2, 
        "Adherence Effect" = 3
    ), 
    bold = TRUE
  ) %>% 
  kable_styling(latex_options = c("scale_down", "HOLD_position"))
```

```{r}
# create frequentist hard drug use and adherence table

# Collect frequentist models in parallel lists matching the outcomes
fmods_base <- list(fmod_ment, fmod_phys, fmod_leu3n, fmod_vload)
fmods_adh  <- list(fmod_ment_adh, fmod_phys_adh, fmod_leu3n_adh, fmod_vload_adh)

row_labels <- c(
  "SF-36 MCS",
  "SF-36 PCS",
  "CD4+ T Cell Count",
  "Log Viral Load"
)

table_adh_comparison_freq <- pmap_dfr(
  list(row_labels, fmods_base, fmods_adh),
  function(outcome_label, fit_base, fit_adh) {
    
    # 1. Unadjusted Hard drugs effect (Base model)
    res_no_adh <- extract_lm_stats(fit_base, "hard_drugs_baseline")
    
    # 2. Adjusted Hard drugs effect (Model with adherence)
    res_adh <- extract_lm_stats(fit_adh, "hard_drugs_baseline")
    
    # 3. Adherence effect (Model with adherence)
    res_adh_term <- extract_lm_stats(fit_adh, "adh")
    
    # 4. Format p-value for the adherence estimate
    p_str <- if (res_adh_term$p < 0.001) "<0.001" else fmt(res_adh_term$p, digits = 3)
    
    tibble::tibble(
      Outcome = outcome_label,
      `Unadj_Est` = fmt(res_no_adh$estimate, 2),
      `Unadj_CI` = sprintf("[%s, %s]", fmt(res_no_adh$lcl, 2), fmt(res_no_adh$ucl, 2)), 
      `Adj_Est` = fmt(res_adh$estimate, 2),
      `Adj_CI` = sprintf("[%s, %s]", fmt(res_adh$lcl, 2), fmt(res_adh$ucl, 2)), 
      `adh_est` = fmt(res_adh_term$estimate, 2),
      `adh_CI  ` = sprintf("[%s, %s]", fmt(res_adh_term$lcl, 2), fmt(res_adh_term$ucl, 2)),
      `p_value` = p_str
    )
  }
) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    linesep = "",
    caption = "Frequentist Estimates: Impact of Adjusting for Adherence on the Effect of Baseline Hard Drug Use (HDU)",
    escape = FALSE,
    align = "lcccccccc", # 8 columns total
    col.names = c(
      "Outcome", 
      "Estimate", 
      "95\\% CI", 
      "Estimate", 
      "95\\% CI", 
      "Estimate", 
      "95\\% CI", 
      "p-value"
    )
  ) %>%
  add_header_above(
    c(
        " " = 1,
        "Unadjusted HDU Effect" = 2, 
        "Adjusted HDU Effect" = 2, 
        "Adherence Effect" = 3
    ), 
    bold = TRUE
  ) %>% 
  kable_styling(latex_options = c("scale_down", "HOLD_position"))
```

```{r, eval=FALSE}
# diagnostics for Bayesian models
bmods <- list(bmod_ment, bmod_phys, bmod_leu3n, bmod_vload,
              bmod_ment_red, bmod_phys_red, bmod_leu3n_red, bmod_vload_red,
              bmod_ment_adh, bmod_phys_adh, bmod_leu3n_adh, bmod_vload_adh)

bmod <- bmod_vload_adh
summary(bmod) # check R-hat (~1) and ESS (>1000)
mcmc_plot(bmod, type = "dens_overlay") # all chains should match
mcmc_plot(bmod, type = "trace") # check for good mixing
mcmc_plot(bmod, type = "acf") # check for lower autocorrelation
```

```{r}
# create histograms for each treatment response by drug use
figure_differences <- hiv_analysis %>%
  pivot_longer(cols = starts_with("diff"), names_to = "outcome", values_to = "change") %>%
  # Recode the outcome variable for clean facet labels and set the order
  mutate(
    outcome = factor(
      outcome, 
      levels = c("diff_agg_ment", "diff_agg_phys", "diff_leu3n", "diff_log_vload"),
      labels = c("SF-36 MCS", "SF-36 PCS", "CD4+ T Cell Count", "Log10 Viral Load")
    )
  ) %>%
  ggplot(aes(x = hard_drugs_baseline, y = change, fill = hard_drugs_baseline)) +
  # Add a dashed reference line at 0 
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.8) +
  # Hide the redundant legend and add slight transparency to the fill
  geom_boxplot(na.rm = TRUE, show.legend = FALSE, alpha = 0.8) +
  facet_wrap(~outcome, scales = "free_y") +
  # Use a clean, publication-standard theme
  theme_bw(base_size = 12) + 
  labs(
    # title = "Treatment Response (Year 0 to 2) by Baseline Hard Drug Use",
    x = NULL,
    y = "Change from Baseline"
  ) +
  theme(
    # Clean up the facet headers
    strip.background = element_rect(fill = "gray95", color = "gray30"),
    strip.text = element_text(face = "bold"),
    # Remove minor grid lines for a less cluttered look
    panel.grid.minor = element_blank(),
    # Increase x-axis text size slightly for readability
    axis.text.x = element_text(size = 11, color = "black"),
    plot.title = element_text(face = "bold")
  )
```


# Introduction

HIV (human immunodeficiency virus) is a virus that attacks the immune system; it destroys T-cells and inhibits the body's ability to fight infection. This analysis evaluates the effectiveness of highly active antiretroviral treatment (HAART), the standard treatment for HIV patients. The data come from the Multicenter AIDS Cohort Study, an ongoing study investigating HIV infection in homosexual and bisexual men in the United States. The dataset includes longitudinal demographic information, laboratory measurements, and 36-Item Short Form Health Survey (SF-36) quality of life summaries on men infected with HIV. Data was collected on subjects prior to beginning HAART (baseline) and every year thereafter for up to 8 years. The primary goal of this analysis is to characterize how treatment response after 2 years of HAART differs based on a subjects hard drug use status at baseline. Treatment response is evaluated across four primary measures: SF-36 Mental Component Summary (MCS) score, SF-36 Physical Component Summary (PCS) score, HIV viral load, and CD4+ T cell count. Specifically, we will model the difference in each primary measure between baseline and Year 2 across hard drug use status, adjusting for baseline measurements, age, BMI, smoking status, education, and race. Additionally, we will investigate how adjusting for study protocol adherence impacts the estimated effects of hard drug use.

# Methods

## Data Management

This analysis only considered subjects with the four primary measures---PCS, MCS, CD4+ count, and viral load---measured before starting HAART (baseline) and after two years of treatment (Year 2). Consistent with standard practice in the literature, viral load was analyzed on the $\log_{10}$ scale, where a one-unit increase corresponds to a 10-fold (order of magnitude) increase in viral load. For this analysis, all demographic variables were measured at baseline and adherence was measured at Year 2. Due to low sample sizes within specific strata, the categories of several demographic variables were combined: race/ethnicity was combined into "Non-Hispanic White" and "Other", education was combined into "No College Degree" and "College Degree or Greater", and smoking was combined into "Current Smokers" and "Never/Former Smokers". Study protocol adherence was dichotomized as $\ge95\%$ and $<95\%$. Additionally, biologically improbable BMI measurements (BMI $<0$ or $>400$) were excluded from the analysis. We investigated if the proportion of missingness in the primary measures differed between hard drug users and non-hard drug users. Finally, for each primary measure, the treatment response (Year 2 minus baseline) was calculated and assessed for normality and outliers. A table of descriptive statistics was created for the demographic variables, adherence, and primary measures at baseline.

## Analysis Approach

Clinically meaningful treatment responses were defined as a 2-unit change in MCS/PCS scores, a 50-cell/mm³ change in CD4+ count, and a 0.5 log10 change in viral load. Following descriptive visualization via boxplots, we fit frequentist and Bayesian multiple linear regression models with normal likelihoods for each treatment response. Models evaluated baseline hard drug use as the primary predictor, adjusting for baseline outcome values, BMI, age, race, education, and smoking status. Results were summarized using point estimates, 95% CIs, and p-values (frequentist), alongside 95% HDIs, posterior probabilities, and LOO-IC (Bayesian). A 3-unit change in LOO-IC was defined as meaningful. To assess the role of adherence on the hard drug use effect, we refit all models to include an adherence term and compared the hard drug use estimates between models with and without this term.

All frequetist analyses were performed using a type one error rate of ($\alpha = 0.05$). All models and figures were created with R version 4.5.1 and Bayesian modeling used `brms` version 2.23.0. ChatGPT version 5.2 was used to debug code, outline the report, and copy-edit the report.

## Bayesian Setup and Diagnositics

For all Bayesian models, the No-U-Turn Sampler (NUTS) was used with 4 chains. Parameters were initialized using random values sampled from a $\operatorname{Unif}(-2, 2)$ distribution on the unconstrained parameter space. Each chain ran for 1000 warmup iterations and 2000 total iterations. The Bayesian models employed non-informative priors: intercepts and fixed effects had $\operatorname{Normal}(0, 100)$ priors, and the residual standard errors had $\operatorname{Half-Normal}(0, 100)$ priors.

The same diagnostic procedure was followed for all of the Bayesian models. First, we ensured the effective sample size (ESS) was sufficient for all parameters (> 1000) and that the R-hat (Gelman-Rubin statistic) was close to 1, indicating convergence. Convergence was further assessed by plotting each parameter's density and comparing between chains. Additionally, trace and autocorrelation plots were visually inspected to evaluate chain mixing and sample dependency.

# Results

Of the 715 subjects in the dataset, 476 had all four primary measures recorded at both baseline and Year 2. These 476 subjects comprised the initial analysis dataset. Notably, missingness in the primary measures was higher among the hard drug use group (41%) compared to the non-hard drug use group (33%), suggesting that missingness may be dependent on baseline hard drug use status. *Table 1* presents the baseline descriptive statistics for all demographic variables and primary measures. Within this initial cohort, 13 subjects had missing or biologically improbable BMI values and were excluded, resulting in a final complete-case analysis dataset of 463 subjects. All Bayesian models had R-hat values close to 1, effective sample sizes over 1000, and no indication of issues with convergence, chain mixing, or sample dependency.

```{r, include=TRUE}
table1
```


## Hard Drug Use Effects

*Figure 1* visualizes the treatment response across hard drug use for each primary measurement. The model results for the hard drug use effect are shown in *Table 2*. For the SF-36 MCS score, there was no significant difference in treatment response between hard drug users and non-users (Frequentist Estimate: -0.29, 95% CI: [-3.75, 3.17], p = 0.870). The Bayesian results were consistent, showing an estimated difference of -0.30 (95% HDI: [-3.57, 3.32]), with a low probability (0.272) of a clinically meaningful effect and no improvement in model fit ($\Delta$ LOO-IC = -3.51). In contrast, hard drug users had a significantly lower treatment response in SF-36 PCS scores compared to non-users (Frequentist Estimate: -3.34, 95% CI: [-6.06, -0.61], p = 0.017). Bayesian estimates similarly showed a -3.33 unit difference (95% HDI: [-5.99, -0.66]), with a 0.834 probability of a clinically significant effect and a meaningful improvement in model fit ($\Delta$ LOO-IC = 3.19). For CD4+ Count, hard drug users experienced a significantly lower response, estimated with a frequentist approach at -163.88 cells/mm³ (95% CI: [-226.35, -101.41], p < 0.001). The Bayesian model estimated a similar, but somewhat smaller, reduction of -148.99 cells/mm³ (95% HDI: [-206.12, -85.85]), with near certainty (probability = 0.999) of a clinically significant effect and a substantially better model fit ($\Delta$ LOO-IC = 24.31). Finally, there was no significant difference in log10 viral load treatment response between the groups (Frequentist Estimate: -0.06, 95% CI: [-0.46, 0.34], p = 0.770). The Bayesian results aligned, estimating a -0.05 difference (95% HDI: [-0.44, 0.33]), a very low probability (0.018) of clinical significance, and no improvement in model fit ($\Delta$ LOO-IC = -1.94).

```{r, include=TRUE, fig.cap="Treatment Responses Across Hard Drug Use Status"}
figure_differences
```

```{r, include=TRUE}
table_freq_bayes
```

## Effects of Adjusting for Adherence

As shown in *Table 1*, hard drug users in the analysis dataset exhibited a higher rate of study protocol adherence compared to non-users (97% vs. 89% achieving $\ge 95\%$ adherence). Notably, only a single subject within the hard drug use cohort fell into the $<95\%$ adherence category. Due to this extreme lack of diversity in adherence among hard drug users, the data provides insufficient information to evaluate adherence as a potential confounder or mediator. Consequently, the inclusion of an adherence term (*Table 3* and *Table 4*) did not meaningfully change the estimated effects of baseline hard drug use. For the outcomes significantly associated with hard drug use---SF-36 PCS and CD4+ count---the estimates changed in magnitude by only 7.2% and 3.5%, respectively, while the non-significant effects remained near zero. Both the Bayesian and frequentist approaches produced consistent results for this part of the analysis.

```{r, include=TRUE}
table_adh_comparison_bayes
```

```{r, include=TRUE}
table_adh_comparison_freq
```

# Discussion and Conclusions

There was not evidence of a statistically or clinically significant impact of hard drug use on the treatment response for viral load or mental quality of life (SF-36 MCS). The treatment response for SF-36 PCS scores was lower by about 3.3 units for subjects who used hard drugs, indicating that HAART was less effective at improving physical quality of life among hard drug users compared to non-users. However, there was not strong evidence that this effect met the 2-unit threshold for clinical significance (95% CI: [-6.06, -0.61], posterior probability: 83.4%). Hard drug use severely diminished CD4+ T cell count response, with both statistical frameworks estimating a reduction of approximately three times the 50 cell/mm³ clinical threshold (estimated at -164 and -149 cells/mm³ for the frequentist and Bayesian models, respectively).

There was no evidence that study protocol adherence acts as a mediator or confounder in the relationship between hard drug use and treatment response across any of the primary measures. However, this assessment is severely limited by the lack of subjects with under 95% adherence in the hard drug use group. Additionally, the higher rate of missing data among hard drug users points to potential attrition bias; subjects who used hard drugs and struggled with adherence may have simplydropped out of the study before their Year 2 measurements were recorded.

Overall, the frequentist and Bayesian approaches produced consistent results, leading to identical conclusions across all four primary measures. The only notable difference occurred in the CD4+ count estimates, likely a consequence of the prior penalizing the exceptionally large scale of the outcome. Future models evaluating CD4+ counts should consider utilizing a flatter prior to mitigate this shrinkage.

# Reproducible Research

## GitHub Link

`https://github.com/nageljo/BIOS6624`

## Dataset Filepath

`C:\Users\josep\OneDrive\AdvancedData\BIOS6624\Project1\Data\hiv_6624_final.csv`